### 2025-03-13 writing/mean-variance-and-scale-of-ndvi-and-the-beta-distribution.Rmd
started appendix for mu and s2 of NDVI and beta

---
title: "Mean and variance of NDVI and the Beta distribution"
author: "Stefano Mezzini"
date: "2025-03-13"
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newcommand*\e{\text{E}} <!-- expectation; non-italic E -->
\newcommand*\var{\text{Var}} <!-- variance; non-italic Var -->
\newcommand*\ra{\text{R}} <!-- range; non-italic R -->

Since NDVI ranges $\nu \in [-1, 1] \approx (-1, 1)$, we can apply the linear transformation $\nu_*=\frac{\nu+1}{2}$ so that $\nu_* \in (0, 1),$ and $\nu_* \sim \text{Beta}(a_*, b_*)$. We can re-parameterize the distribution as $\nu_* \sim \text{Beta}(\mu_*, \phi_*)$ in terms of the mean $$\mu_* = \frac{a_*}{a_*+b_*}$$ and scale $$\phi_* = \frac{\sigma^2_*}{\mu_*~(1 - \mu_*)} = \frac{a_*b_*}{(a_*+b_*)^2(a_*+b_*+1)} \frac{a_*+b_*}{a_*} \frac{1}{\frac{a_* + b_* - a_*}{a_*+b_*}} = \frac{1}{(a_*+b_*+1)}.$$ For each parameter, we have the ranges
$$\ra(\mu_*) = \left(\frac{0}{0+\infty}, \frac{\infty}{\infty+0}\right) = (0, 1),$$
$$\ra(\sigma^2_*) = \left(\frac{0}{(0)^2(0+0+1)}, \frac{c^2}{4c^2(2c+1)}\right) = \left(\frac{0}{(0)^2(0+0+1)}, \frac{1}{8(0^+)+4}\right) = (0, 0.25),$$
and $$\ra(\phi_*) \left(\frac{1}{(\infty+\infty+1)}, \frac{1}{(0+0+1)}\right) = (0, 1).$$

For NDVI, the mean is $$\mu = \e(\nu) = \e(2 \nu_* - 1) = 2~\e(\nu_*) - 1 = 2\mu_* - 1,$$ the variance is $$\sigma^2 = \var(\nu) = \var(2 \nu_* - 1) = 4~\var(\nu_*) = 4\sigma^2_*,$$ and the ranges for $\mu$ and $\sigma^2$ are $\ra(\mu) = (-1, 1)$ and $\ra(\sigma^2) = (0, 1)$, respectively. Calculating the scale parameter for NDVI is not necessary because our focus is on the mean and variance of NDVI.

### 2025-03-13 analysis/000-sardinia-test-no-mrf.R
added mrf to sardinia test

@@ -1,223 +1,0 @@
-library('sf')
-library('terra')
-library('elevatr')
-library('dplyr')
-library('lubridate')
-library('purrr')
-library('furrr')
-library('mgcv')
-library('ggplot2')
-library('gratia')
-source('functions/betals.r') # custom beta location-scale family
-source('functions/scale-ndvi.R')
-source('functions/ndvi-palette.R')
-source('analysis/figures/default-ggplot-theme.R')
-theme_set(theme_bw())
-
-# the bounding box for sardinia (large enough to include all coast)
-sardinia_bbox <- tibble(x = c(7.6, 10.4), y = c(38.8, 41.3)) %>%
-  st_as_sf(coords = c('x', 'y')) %>%
-  st_bbox() %>%
-  st_as_sfc() %>%
-  st_as_sf() %>%
-  st_set_crs('EPSG:4326')
-
-# shapefile of the world's terrestrial ecoregions
-sardinia <- read_sf('data/world-ecosystems/data/commondata/data0/tnc_terr_ecoregions.shp') %>%
-  st_geometry() %>%
-  st_make_valid() %>%
-  st_intersection(sardinia_bbox)
-
-# import ndvi data (downloaded from the NASA website)
-if(file.exists('data/sardinia-test/sardinia-ndvi.rds')) {
-  sardinia_ndvi <- readRDS('data/sardinia-test/sardinia-ndvi.rds')
-} else {
-  if(.Platform$OS.type != 'unix') stop('NOAA rasters are on the lab Linux.')
-  future::availableCores(logical = FALSE)
-  plan(multisession, workers = 30)
-  sardinia_ndvi <-
-    list.files(path = '/home/shared/NOAA_Files/',
-               pattern = '.nc',
-               full.names = TRUE) %>%
-    future_map(\(fn) {
-      r <- rast(fn)
-      r %>%
-        crop(sardinia_bbox) %>% # to avoid including the rest of italy
-        mask(sardinia_bbox) %>%
-        as.data.frame(xy = TRUE, na.rm = TRUE) %>%
-        mutate(date = as.Date(unique(time(r))))
-    }, .progress = TRUE) %>%
-    bind_rows() %>%
-    as_tibble() %>%
-    rename(ndvi = NDVI) %>%
-    mutate(ndvi_scaled = ndvi_to_01(ndvi))
-  plan(sequential)
-  saveRDS(sardinia_ndvi, 'data/sardinia-test/sardinia-ndvi.rds')
-}
-
-if(FALSE) {
[…]

### 2025-03-13 analysis/000-sardinia-test.R

@@ -1,0 +1,451 @@
+# all tests ran on my personal laptop:
+# 64.0 GB RAM, 13th Gen Intel Core i7-1370P processor (1.90 GHz)
+library('sf')
+library('terra')
+library('elevatr')
+library('dplyr')
+library('lubridate')
+library('purrr')
+library('furrr')
+library('mgcv')
+library('ggplot2')
+library('gratia')
+source('functions/betals.r') # custom beta location-scale family
+source('functions/scale-ndvi.R')
+source('functions/ndvi-palette.R')
+source('analysis/figures/default-ggplot-theme.R')
+theme_set(theme_bw())
+
+# the bounding box for sardinia (large enough to include all coast)
+sardinia_bbox <- tibble(x = c(7.6, 10.4), y = c(38.8, 41.3)) %>%
+  st_as_sf(coords = c('x', 'y')) %>%
+  st_bbox() %>%
+  st_as_sfc() %>%
+  st_as_sf() %>%
+  st_set_crs('EPSG:4326')
+
+# shapefile of the world's terrestrial ecoregions
+sardinia <- read_sf('data/world-ecosystems/data/commondata/data0/tnc_terr_ecoregions.shp') %>%
+  st_make_valid() %>%
+  st_intersection(sardinia_bbox) %>%
+  st_cast('POLYGON') %>%
+  st_as_sf() %>%
+  mutate(poly_id = paste('poly', 1:n()),
+         area_km2 = as.numeric(st_area(.)) / 1e6)
+hist(sardinia$area_km2)
+
+# import ndvi data (downloaded from the NASA website) ----
+if(file.exists('data/sardinia-test/sardinia-ndvi.rds')) {
+  sardinia_ndvi <- readRDS('data/sardinia-test/sardinia-ndvi.rds')
+} else {
+  if(.Platform$OS.type != 'unix') stop('NOAA rasters are on the lab Linux.')
+  future::availableCores(logical = FALSE)
+  plan(multisession, workers = 30)
+  sardinia_ndvi <-
+    list.files(path = '/home/shared/NOAA_Files/',
+               pattern = '.nc',
+               full.names = TRUE) %>%
+    future_map(\(fn) {
+      r <- rast(fn)
+      r %>%
+        crop(sardinia_bbox) %>% # to avoid including the rest of italy
+        mask(sardinia_bbox) %>%
+        as.data.frame(xy = TRUE, na.rm = TRUE) %>%
[…]

### 2025-03-13 analysis/001-ecoregion-neighbors.R
fixed typo in comment

@@ -76,7 +76,7 @@ length(nbs) == nrow(ecoregions)
 # add names corresponding to each polygon
 all(attr(nbs, 'region.id') == ecoregions$poly_id)
 names(nbs) # no names, currently
-names(nbs) <- attr(nbs, 'region.id') # names do not need to match indeces
+names(nbs) <- attr(nbs, 'region.id') # names do not need to match indices
 
 # add number of neighbors for each polygon
 ecoregions$n_neigh <- map_int(nbs, function(.nb) {

### 2025-03-31 analysis/figures/default-ggplot-theme.R
removed unnecessary `plot_scheme()` call

@@ -8,7 +8,7 @@ theme_set(theme_bw() +
                   text = element_text(face = 'bold')))
 
 # custom NDVI color palette
-ndvi_pal <- plot_scheme(color('bukavu')(30)) # wrong order
+ndvi_pal <- color('bukavu')(30) # wrong order
 if(.PLOT_PALETTES) plot_scheme(ndvi_pal)
 ndvi_pal <- ndvi_pal[c(1:13, length(ndvi_pal):(length(ndvi_pal)/2 + 1))]
 if(.PLOT_PALETTES) plot_scheme(ndvi_pal)
@@ -35,4 +35,3 @@ if(.PLOT_PALETTES) {
      scale_fill_gradientn(colours = ndvi_pal, limits = c(-1, 1))) %>%
     colorblindr::cvd_grid()
 }
-

### 2025-03-31 analysis/000-sardinia-test.R
added test for `ti` term

@@ -449,3 +449,30 @@ cowplot::plot_grid(
 ggsave(paste0('figures/sardinia-test/sardinia-ndvi-model-agreement-duchon-mrf-',
               unique(preds_1_mrf$date), 'predictions.png'),
        height = 12, width = 8, bg = 'white')
+
+# testing complexity of ti(doy, space) ----
+#' `ti()` of doy and space doesn't add much
+# on personal laptop: "initial" is 50 s, run is 5 s
+if(file.exists('models/sardinia-test/gaus-gam-ti-ds-gam.rds')) {
+  m_gaus_ti_ds <- readRDS('models/sardinia-test/gaus-gam-ti-ds-gam.rds')
+} else {
+  m_gaus_ti_ds <- bam(ndvi_scaled ~
+                      s(x, y, bs = 'ds', k = 200) +
+                      s(elev_m, bs = 'ad', k = 10) + # adaptive spline to correct for shorelines better
+                      s(year, bs = 'cr', k = 10) +
+                      s(doy, bs = 'cc', k = 10) +
+                      ti(doy, year, bs = c('cc', 'cr'), k = c(5, 5)),
+                    family = gaussian(),
+                    knots = list(doy = c(0, 1)),
+                    data = sardinia_ndvi,
+                    method = 'fREML',
+                    discrete = TRUE,
+                    control = gam.control(nthreads = 1, trace = TRUE))
+  saveRDS(m_gaus_ti_ds, 'models/sardinia-test/gaus-gam-ti-ds-gam.rds')
+}
+
+draw(m_gaus_ti_ds, rug = FALSE)
+ggsave('figures/sardinia-test/sardinia-ndvi-gaussian-ti-terms.png',
+       width = 13.5, height = 6, units = 'in', dpi = 300, bg = 'white')
+
+summary(m_gaus_ti_ds)

### 2025-03-31 analysis/figures/input-data.R
created maps of input data

@@ -1,0 +1,91 @@
+library('dplyr')     # for data wrangling
+library('sf')        # for simple features
+library('terra')     # for rasters
+library('ggplot2')   # for fancy plots
+library('tidyterra') # for plotting spatRasters with ggplot2
+library('khroma')    # for colorblind-friendly palettes
+source('analysis/figures/default-ggplot-theme.R')
+
+eco <- read_sf('data/ecoregions/ecoregions-polygons.shp') %>%
+  filter(WWF_MHTNAM != 'Inland Water') %>%
+  mutate()
+
+if(FALSE) {
+khroma::info() %>%
+  filter(type == 'qualitative') %>%
+  pull(palette)
+  
+  # potential discrete palettes
+  plot_scheme(color('bright')(7))
+  plot_scheme(color('highcontrast')(3))
+  plot_scheme(color('vibrant')(7))
+  plot_scheme(color('discreterainbow')(16))
+  
+  khroma::info() %>%
+    filter(type == 'sequential') %>%
+    pull(palette)
+}
+
+# individual polygons ----
+n_distinct(eco$poly_id)
+pal_polys <- c(color('okabeito')(8), 'grey')
+plot_scheme_colorblind(pal_polys)
+
+p_poly <-
+  ggplot(eco) +
+  geom_sf(fill = pal_polys[1:nrow(eco) %% length(pal_polys) + 1],
+          lwd = .05) +
+  scale_x_continuous(expand = c(0, 0))
+
+ggsave('figures/input-data/polygons.png', p_poly,
+       width = 10, height = 4.4, units = 'in', dpi = 600, bg = 'white')
+
+# ecoregions ----
+n_distinct(eco$WWF_MHTNAM)
+
+p_eco <-
+  ggplot(eco) +
+  geom_sf(aes(fill = WWF_MHTNAM), color = 'black', lwd = .05) +
+  scale_fill_discreterainbow(name = 'Ecoregion') +
+  scale_x_continuous(expand = c(0, 0)) +
+  theme(legend.position = 'top', legend.text = element_text(size = 4.5))
+ggsave('figures/input-data/ecoregions.png', p_eco,
+       width = 10.1, height = 6, units = 'in', dpi = 600, bg = 'white')
+
+# elevation ----
+r_elev <- rast('data/elev-raster.tif') %>%
+  crop(., st_transform(eco, crs(.)), mask = TRUE)
+
+hist(r_elev)
[…]

### 2025-03-31 writing/manuscript.Rmd
created manuscript doc

---
title: \large A new global measure of environmental unpredictability 
author:
  - name: Stefano Mezzini
    institute: [braes, biol]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
    correspondence: "yes"
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'ndvi-stochasticity.bib'
csl: 'freshwater-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
    keep_tex: false # for submitting TeX file
editor_options:
  chunk_output_type: console
---

```{r, echo=FALSE}
abstract <- ""
```

<!-- elements of the title page -->

\clearpage

\noindent \textbf{Article type}: Research article

\noindent \textbf{Words in abstract}: `r stringi::stri_count_words(abstract)`

\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count())`

\noindent \textbf{Figures}: \textcolor{red}{\textbf{XXX}}

\noindent \textbf{Tables}: \textcolor{red}{\textbf{XXX}}

\noindent \textbf{References}: \textcolor{red}{\textbf{XXX}}

\noindent \textbf{Appendices}: \textcolor{red}{\textbf{XXX}}

\noindent \textbf{Key words:} remote sensing, NDVI, environmental stochasticity, \textcolor{red}{\textbf{XXX}}

<!-- start main body on a new page -->

\newpage

\doublespacing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '100%', fig.align = 'center')
```

\linenumbers

# Abstract

\noindent `r abstract`

\newpage

# Introduction

https://www.sciencedirect.com/science/article/abs/pii/S0034425718305625
https://silvis.forest.wisc.edu/data/dhis/

# Methods

maps of input data

NDVI palette is colorblid-friendly

# Results

output of `draw()`

hex plots of mean and var vs common metrics (e.g., spp diversity, npp, hfi, ...)

# Discussion

# Useful references

* `{khroma}`: doi:10.5281/zenodo.1472077
* @keith_predicting_2008
* @chevin_adaptation_2010
* @rickbeil_plasticity_2019
* @mueller_search_2008
* @pettorelli_using_2005
* @keys_machine-learning_2021
* @nilsen_can_2005
* @stephens_optimal_1982
* @merkle_large_2016
* @tian_evaluating_2015
* @huang_commentary_2021
* @fan_global_2016
* @wang_stochastic_2019
* @pease_ecological_2024
* @xu_plasticity_2021
* @mezzini_how_2025

# References

### 2025-03-31 analysis/000-sardinia-test.R
added `st_geometr()` when plotting over raster

@@ -75,7 +75,7 @@ if(file.exists('data/sardinia-test/sardinia-ndvi.rds')) {
     rast(crs = 'EPSG:4326') %>%
     get_elev_raster(z = 6)
   plot(elevs)
-  plot(sardinia, add = TRUE, col = 'transparent')
+  plot(st_geometry(sardinia), add = TRUE, col = 'transparent')
   
   sardinia_ndvi <- mutate(sardinia_ndvi, 
                           elev_m = extract(elevs, select(sardinia_ndvi, x, y)),
@@ -142,7 +142,7 @@ ggsave('figures/sardinia-test/sardinia-ndvi-beta-terms-ndvi-scale.png',
 summary(m_beta)
 
 # fit a spatially explicit test model with a betals family ----
-#' fails with error (*tested on lab' linux machine*):
+#' fails with error (*tested on lab linux machine*):
 #'`Error in h(simpleError(msg, call)) :`
 #'`error in evaluating the argument 'x' in selecting a method for function 't':`
 #'`long vectors (argument 5) are not supported in .Fortran`

### 2025-03-31 analysis/001-ecoregion-neighbors.R
will create neighbor lists in the modeling scripts

@@ -5,6 +5,7 @@ library('sf')      # for spatial data
 library('terra')   # for rasters
 library('spdep')   # for finding neighbors
 library('ggplot2') # for fancy plots
+source('functions/add_nb.R')
 
 # get shapefile of each continent
 ecoregions <- st_read('data/world-ecosystems/data/commondata/data0/tnc_terr_ecoregions.shp') %>%
@@ -31,7 +32,8 @@ ecoregions <- st_read('data/world-ecosystems/data/commondata/data0/tnc_terr_ecor
   st_cast('POLYGON') %>% # assigns attributes to sub-geometries by default
   st_make_valid() %>% # to drop duplicate vertices
   mutate(poly_id = paste('poly', 1:n()),
-         area_km2 = as.numeric(st_area(.)) / 1e6)
+         area_km2 = as.numeric(st_area(.)) / 1e6) %>%
+  filter(WWF_MHTNAM != 'Inland Water') # dropping 17 inland water polygons
 
 # drop points based on plot(n_data ~ area)
 # drop points with areas < (5 km)^2 * 1, 4, or 9 cells
@@ -77,6 +79,7 @@ length(nbs) == nrow(ecoregions)
 all(attr(nbs, 'region.id') == ecoregions$poly_id)
 names(nbs) # no names, currently
 names(nbs) <- attr(nbs, 'region.id') # names do not need to match indices
+names(nbs)
 
 # add number of neighbors for each polygon
 ecoregions$n_neigh <- map_int(nbs, function(.nb) {
@@ -157,43 +160,6 @@ plot(ru_geom[1:4, ], col = pal)
 plot(ru_geom[5:6, ], col = pal[5:6])
 layout(1)
 
-add_nb <- function(p1, p2, add = FALSE) {
-  if(! all(c(p1, p2) %in% names(nbs))) {
-    stop('At least one of ', p1, ' or ', p2, ' is not in `names(nbs)`.')
-  }
-  
-  # add neighbors if polygons are not already neighbors
-  if(which(names(nbs) == p1) %in% nbs[[p2]] |
-     which(names(nbs) == p2) %in% nbs[[p1]]) {
-    stop('Polygons ', p1, ' and ', p2,
-         ' are already neighbors! Check the values for each.')
-  } else {
-    filter(ecoregions, poly_id == p1 | poly_id == p2) %>%
-      st_geometry() %>%
[…]

### 2025-03-31 functions/add_nb.R

@@ -1,0 +1,39 @@
+library('dplyr')
+library('sf')
+
+add_nb <- function(p1, p2, add = FALSE) {
+  if(! all(c(p1, p2) %in% names(nbs))) {
+    stop('At least one of ', p1, ' or ', p2, ' is not in `names(nbs)`.')
+  }
+  
+  # add neighbors if polygons are not already neighbors
+  if(which(names(nbs) == p1) %in% nbs[[p2]] |
+     which(names(nbs) == p2) %in% nbs[[p1]]) {
+    stop('Polygons ', p1, ' and ', p2,
+         ' are already neighbors! Check the values for each.')
+  } else {
+    filter(ecoregions, poly_id == p1 | poly_id == p2) %>%
+      st_geometry() %>%
+      st_transform('ESRI:102007') %>% # hawai'i albers
+      plot(col = 1:2, main = paste('Polygons', p1, 'and', p2))
+    
+    if(add) {
+      if(all(nbs[[p1]] == 0)) { # if no other neighbors do not add "0"
+        nbs[[p1]] <<- which(names(nbs) == p2)
+      } else { # otherwise keep other neighbors
+        nbs[[p1]] <<- c(nbs[[p1]], which(names(nbs) == p2))
+      }
+      
+      if(all(nbs[[p2]] == 0)) { # if no other neighbors do not add "0"
+        nbs[[p2]] <<- which(names(nbs) == p1)
+      } else { # otherwise keep other neighbors
+        nbs[[p2]] <<- c(nbs[[p2]], which(names(nbs) == p1))
+      }
+      
+      cat('Added', p1, 'as a neighbor to', p2, 'and vice-versa.\n')
+    } else {
+      warning('Did not add ', p1, ' as a neighbor to ', p2,
+              ' and vice-versa!')
+    }
+  }
+}

### 2025-03-31 analysis/002-download-ndvi-rasters.R
added check for corrupt rasters

@@ -95,4 +95,12 @@ length(fn_dates) == length(main_dates)
 length(fn_dates) - length(main_dates)
 as.numeric(diff(range(main_dates), units = 'days')) / 365
 
-#' *need to move files to "//home/shared" folder on the lab Linux*
+# ensure no rasters are corrupt
+library('terra')
+file_names <- list.files(path = 'data/avhrr-viirs-ndvi/raster-files/',
+                         pattern = '.nc', full.names = TRUE, recursive = FALSE)
+table(map_chr(file_names, function(.fn) class(rast(.fn)), .progress = TRUE))
+
+# if any are corrupt, download manually from
+# https://www.ncei.noaa.gov/data/land-normalized-difference-vegetation-index/access/
+#' *need to check if all files are in "//home/shared" folder on the lab Linux*

### 2025-03-31 analysis/003-merging-rasters.R
merged rasters; made subsets for 2020 and 2021

@@ -32,6 +32,7 @@ if(FALSE) {
 ecoregions <- st_read('data/world-ecosystems/data/commondata/data0/tnc_terr_ecoregions.shp') %>%
   filter(WWF_REALM2 != 'Antarctic') # drop Antarctica
 
+# all rasters have the same CRS
 # check crs for first, last, and a random sample of rasters
 c(file_names[1], sample(file_names, size = 100), file_names[length(file_names)]) %>%
   map_chr(function(.fn) { # fast enough that it is not worth parallelizing
@@ -39,7 +40,9 @@ c(file_names[1], sample(file_names, size = 100), file_names[length(file_names)])
   }) %>%
   unique()
 
-ecoregions <- st_transform(ecoregions, crs(rast(file_names[1])))
+ecoregions <- st_transform(ecoregions, crs(rast(file_names[1]))) %>%
+  st_geometry() %>%
+  vect()
 
 # create a data frame of all dates
 dates <-
@@ -72,38 +75,127 @@ dates %>%
 # some dates are missing a raster (not available on the server)
 all(! is.na(dates$file_name))
 mean(! is.na(dates$file_name))
+filter(dates, is.na(file_name))
 
-# need to aggregate temporally to reduce file size before saving
+# need to aggregate temporally to reduce file size before saving ----
 # spatRast objects cannot be serialized (i.e., run in parallel): https://stackoverflow.com/questions/67445883/terra-package-returns-error-when-try-to-run-parallel-operations/67449818#67449818
+# data for 2020 omly ----
+tictoc::tic() # ~8 minutes on EME Linux
+dates_aggr_2020 <-
+  dates %>%
+  filter(! is.na(file_name)) %>% # drop missing rasters
+  filter(lubridate::year(date) == 2020) %>%
+  nest(cluster = ! group) %>%
+  mutate(central_date = map_dbl(cluster, function(.cl) {
+    mean(.cl$date)
+  }, .progress = 'Getting mean dates') %>%
+    as.Date(), # convert numeric back to date
+  ndvi_rast = map(cluster, function(.cl) {
+    map(.cl$file_name, \(.fn) {
+      rast(.fn, lyr = 'NDVI')
+    }) %>%
+      rast() %>%
+      mask(ecoregions) %>% # drop antarctica and unnecessary pixels
+      mean(na.rm = TRUE) %>%
[…]

### 2025-03-31 analysis/figures/input-data.R
created figure of number of rasters over time

@@ -11,9 +11,9 @@ eco <- read_sf('data/ecoregions/ecoregions-polygons.shp') %>%
   mutate()
 
 if(FALSE) {
-khroma::info() %>%
-  filter(type == 'qualitative') %>%
-  pull(palette)
+  khroma::info() %>%
+    filter(type == 'qualitative') %>%
+    pull(palette)
   
   # potential discrete palettes
   plot_scheme(color('bright')(7))
@@ -24,6 +24,34 @@ khroma::info() %>%
   khroma::info() %>%
     filter(type == 'sequential') %>%
     pull(palette)
+  
+  plot_scheme(color('devon')(10))
+  plot_scheme(color('lajolla')(10))
+  plot_scheme(color('bamako')(10))
+  plot_scheme(color('davos')(10))
+  plot_scheme(color('bilbao')(10))
+  plot_scheme(color('nuuk')(10))
+  plot_scheme(color('oslo')(10))
+  plot_scheme(color('grayC')(10))
+  plot_scheme(color('hawaii')(10))
+  plot_scheme(color('lapaz')(10))
+  plot_scheme(color('tokyo')(10))
+  plot_scheme(color('buda')(10))
+  plot_scheme(color('acton')(10))
+  plot_scheme(color('turku')(10))
+  plot_scheme(color('imola')(10))
+  plot_scheme(color('batlow')(10))
+  plot_scheme(color('batlowW')(10))
+  plot_scheme(color('batlowK')(10))
+  plot_scheme(color('brocO')(10))
+  plot_scheme(color('corkO')(10))
+  plot_scheme(color('vikO')(10))
+  plot_scheme(color('romaO')(10))
+  plot_scheme(color('bamO')(10))
+  plot_scheme(color('YlOrBr')(10))
+  plot_scheme(color('iridescent')(10))
+  plot_scheme(color('incandescent')(10))
+  plot_scheme(color('smoothrainbow')(10))
 }
 
 # individual polygons ----
@@ -89,3 +117,30 @@ p_dist
 
 ggsave('figures/input-data/dist-km.png', p_dist,
        width = 10, height = 5, units = 'in', dpi = 600, bg = 'white')
+
+# hex plot of rasters over day of year and year
+dates <- tibble(
+  file_name = list.files('H:/GitHub/ndvi-stochasticity/data/avhrr-viirs-ndvi/raster-files',
+                         pattern = '*.nc', recursive = FALSE),
[…]

### 2025-03-31 writing/appendix-1.Rmd

--- 2025-03-31 writing/manuscript.Rmd
@@
 title: \large A new global measure of environmental unpredictability 
+subtitle: Appendix 1
 author:
@@
 
-```{r, echo=FALSE}
-abstract <- ""
+\clearpage
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE, out.width = '100%')
 ```
 
-<!-- elements of the title page -->
+```{r packages, warning=FALSE}
+library('knitr')
+```
+
+\noindent The figures below illustrate some of the data used for fitting the models.
+All color palettes below are available via the `khroma` package [@frerebeau_khroma_2024] for `R` [@r_core_team_r_2024].
+The code for generating the figures is available [on GitHub](https://github.com/QuantitativeEcologyLab/ndvi-stochasticity/blob/main/analysis/figures/input-data.R).
+
+```{r ecoregions, fig.cap="Map of the ecoregions."}
+include_graphics('../figures/input-data/ecoregions.png')
+```
+
+```{r polygons, fig.cap="Map of the polygons for each ecoregion. Each polygon is colored with one of 9 colors based on the polygon names. Note that some neighboring polygons have the same color."}
+include_graphics('../figures/input-data/polygons.png')
+```
+
+```{r elevation, fig.cap="Map of elevation above sea level."}
+include_graphics('../figures/input-data/elev-m.png')
+```
+
+```{r distance-coast, fig.cap="Map of the estimated distance from the nearest coastal edge."}
+include_graphics('../figures/input-data/dist-m.png')
+```
+
+```{r n-rasters, fig.cap="Number of days with a raster within 15-day periods starting on January $1^{\\text{st}}$, for each year."}
+include_graphics('../figures/input-data/n-rasters-time.png')
+```
 
@@
 
-\noindent \textbf{Article type}: Research article
-
-\noindent \textbf{Words in abstract}: `r stringi::stri_count_words(abstract)`
-
-\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count())`
-
-\noindent \textbf{Figures}: \textcolor{red}{\textbf{XXX}}
-
-\noindent \textbf{Tables}: \textcolor{red}{\textbf{XXX}}
-
-\noindent \textbf{References}: \textcolor{red}{\textbf{XXX}}
-
-\noindent \textbf{Appendices}: \textcolor{red}{\textbf{XXX}}
-
-\noindent \textbf{Key words:} remote sensing, NDVI, environmental stochasticity, \textcolor{red}{\textbf{XXX}}
-
-<!-- start main body on a new page -->
-
-\newpage
-
-\doublespacing
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, out.width = '100%', fig.align = 'center')
-```
-
-\linenumbers
-
-# Abstract
-
-\noindent `r abstract`
-
-\newpage
-
-# Introduction
-
-https://www.sciencedirect.com/science/article/abs/pii/S0034425718305625
-https://silvis.forest.wisc.edu/data/dhis/
-
-# Methods
-
-maps of input data
-
-NDVI palette is colorblid-friendly
-
-# Results
-
-output of `draw()`
-
-hex plots of mean and var vs common metrics (e.g., spp diversity, npp, hfi, ...)
-
-# Discussion
-
-# Useful references
-
-* `{khroma}`: doi:10.5281/zenodo.1472077
-* @keith_predicting_2008
-* @chevin_adaptation_2010
-* @rickbeil_plasticity_2019
-* @mueller_search_2008
-* @pettorelli_using_2005
-* @keys_machine-learning_2021
-* @nilsen_can_2005
-* @stephens_optimal_1982
-* @merkle_large_2016
-* @tian_evaluating_2015
-* @huang_commentary_2021
-* @fan_global_2016
-* @wang_stochastic_2019
-* @pease_ecological_2024
-* @xu_plasticity_2021
-* @mezzini_how_2025
-
 # References

### 2025-03-31 functions/scale-ndvi.R
added argument to avoid issues due to boundries

@@ -1,3 +1,34 @@
-ndvi_to_01 <- function(x) (x + 1) / 2
+# constant to move values away from boundaries
+ndvi_to_01 <- function(ndvi, avoid_boundaries = FALSE) {
+  ndvi_scaled <- (ndvi + 1) / 2
+  
+  if(avoid_boundaries) {
+    EPS <- .Machine$double.eps * 100
+    
+    ndvi_scaled <- dplyr::if_else(condition = ndvi_scaled - EPS <= 0,
+                                  true = EPS,
+                                  false = ndvi_scaled)
+    ndvi_scaled <- dplyr::if_else(condition = ndvi_scaled + EPS >= 1,
+                                  true = ndvi_scaled - EPS,
+                                  false = ndvi_scaled)
+  }
+  
+  return(ndvi_scaled)
+}
 
-ndvi_to_11 <- function(x) x * 2 - 1
+ndvi_to_11 <- function(ndvi_scaled, avoided_boundaries = FALSE) {
+  if(avoided_boundaries) {
+    EPS <- .Machine$double.eps * 100
+    
+    ndvi_scaled <- dplyr::if_else(condition = ndvi_scaled == EPS,
+                                  true = 0, # = ndvi_scaled - EPS,
+                                  false = ndvi_scaled)
+    ndvi_scaled <- dplyr::if_else(condition = ndvi_scaled == 1 - EPS,
+                                  true = 1, # ndvi_scaled + EPS,
+                                  false = ndvi_scaled)
+  }
+  
+  ndvi <- ndvi_scaled * 2 - 1
+  
+  return(ndvi)
+}

### 2025-03-31 analysis/010-test-hgams.R
wrote script for testing hgam for mean ndvi

@@ -1,0 +1,244 @@
+library('dplyr')   # for data wrangling
+library('purrr')   # for functional programming
+library('sf')      # for shapefiles
+library('mgcv')    # for Genralized Additive Models
+library('sf')      # for simple fleature objects
+library('terra')   # for rasters
+library('elevatr') # for extracting elevation
+library('gratia')  # for plotting GAMs
+source('analysis/figures/default-ggplot-theme.R')
+source('functions/add_nb.R')
+
+ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
+
+# resolution is (0.05 degrees)^2 = (3 degree minutes)^2
+r_0 <- rast('data/avhrr-viirs-ndvi/raster-files/AVHRR-Land_v005_AVH13C1_NOAA-07_19810624_c20170610041337.nc',
+            lyr = 'QA') %>% # to have a value for each cell irrespective of raster
+  aggregate(2) # because aggregated in the dataset
+values(r_0) <- 1
+plot(r_0)
+res(r_0)
+
+# create raster of ecoregion
+if(file.exists('data/ecoregions/wwf-ecoregions.tif')) {
+  r_eco <- rast('data/ecoregions/wwf-ecoregions.tif')
+} else {
+  r_eco <-
+    ecoregions %>%
+    select(WWF_MHTNAM) %>%
+    vect() %>%
+    rasterize(r_0, field = 'WWF_MHTNAM')
+  plot(r_eco, legend = FALSE)
+  writeRaster(r_eco, 'data/ecoregions/wwf-ecoregions.tif')
+}
+
+# create raster of polygon ID
+if(file.exists('data/ecoregions/ecoregion-polygon-id.tif')) {
+  r_poly_id <- rast('data/ecoregions/ecoregion-polygon-id.tif')
+} else {
+  r_poly_id <-
+    ecoregions %>%
+    select(poly_id) %>%
+    vect() %>%
+    rasterize(r_0, field = 'poly_id')
+  plot(r_poly_id, legend = FALSE)
+  writeRaster(r_poly_id, 'data/ecoregions/ecoregion-polygon-id.tif')
+}
+
+# get raster of elevations
+if(file.exists('data/elev-raster.tif')) {
+  r_elev <- rast('data/elev-raster.tif')
+} else {
+  r_elev <- get_elev_raster(ecoregions, z = 3) #'`z=3` gives `0.076 < res(r_0)`
[…]

### 2025-03-31 analysis/misc/visualizing-mean-and-varinance-in-ndvi.R
adding misc scripts to `analysis/misc`

@@ -1,0 +1,21 @@
+rbeta2 <- function(mu, s2) {
+  alpha <- ((1 - mu) / s2 - 1/mu) * mu^2
+  beta <- alpha * (1/mu - 1)
+  
+  cat('alpha =', alpha, ', beta =', beta)
+  x <- rbeta(1e3, alpha, beta)
+  
+  hist(x, main = bquote(mu == .(mu)~','~sigma^2 == .(s2)),
+       ylab = '', xlab = 'NDVI', xlim = c(0, 1))
+}
+
+layout(matrix(1:4, ncol = 2))
+rbeta2(mu = 0.5, s2 = 0.01)
+rbeta2(mu = 0.5, s2 = 0.25) # note: scale is 0.25 / (0.5 * 0.5) = 1
+rbeta2(mu = 0.9, s2 = 0.02)
+rbeta2(mu = 0.9, s2 = 0.089999) # note: scale is 0.089999 / (0.9 * 0.1) = 0.9999889 ~= 1
+
+rbeta2(mu = 0.5, s2 = 0.002)
+rbeta2(mu = 0.5, s2 = 0.02) # note: scale is 0.25 / (0.5 * 0.5) = 1
+rbeta2(mu = 0.9, s2 = 0.02)
+rbeta2(mu = 0.1, s2 = 0.02) # note: scale is 0.089999 / (0.9 * 0.1) = 0.9999889 ~= 1

### 2025-04-01 analysis/figures/010-input-data.R
reordered ecoregions; numbered figure scripts

@@ -8,7 +8,22 @@ source('analysis/figures/default-ggplot-theme.R')
 
 eco <- read_sf('data/ecoregions/ecoregions-polygons.shp') %>%
   filter(WWF_MHTNAM != 'Inland Water') %>%
-  mutate()
+  mutate(WWF_MHTNAM = factor(WWF_MHTNAM, levels = c(
+    'Rock and Ice',
+    'Tundra',
+    'Boreal Forests/Taiga',
+    'Montane Grasslands and Shrublands',
+    'Temperate Grasslands, Savannas and Shrublands',
+    'Temperate Broadleaf and Mixed Forests',
+    'Temperate Conifer Forests',
+    'Mediterranean Forests, Woodlands and Scrub',
+    'Mangroves',
+    'Tropical and Subtropical Moist Broadleaf Forests',
+    'Tropical and Subtropical Dry Broadleaf Forests',
+    'Tropical and Subtropical Coniferous Forests',
+    'Tropical and Subtropical Grasslands, Savannas and Shrublands',
+    'Flooded Grasslands and Savannas',
+    'Deserts and Xeric Shrublands')))
 
 if(FALSE) {
   khroma::info() %>%
@@ -74,9 +89,11 @@ n_distinct(eco$WWF_MHTNAM)
 p_eco <-
   ggplot(eco) +
   geom_sf(aes(fill = WWF_MHTNAM), color = 'black', lwd = .05) +
+  geom_hline(yintercept = 0, color = 'black', lwd = 0.1, lty = 'dashed') +
   scale_fill_discreterainbow(name = 'Ecoregion') +
   scale_x_continuous(expand = c(0, 0)) +
   theme(legend.position = 'top', legend.text = element_text(size = 4.5))
+
 ggsave('figures/input-data/ecoregions.png', p_eco,
        width = 10.1, height = 6, units = 'in', dpi = 600, bg = 'white')
 
@@ -98,10 +115,11 @@ p_elev <-
   scale_x_continuous(expand = c(0, 0)) +
   theme(legend.position = 'top', legend.text = element_text(size = 4.5))
 p_elev
+
 ggsave('figures/input-data/elev-m.png', p_elev,
        width = 10, height = 5, units = 'in', dpi = 600, bg = 'white')
 
-# distance from coast?
+# distance from coast
 r_dist <- rast('data/distance-from-coast-m.tif') %>%
[…]

### 2025-04-01 analysis/000-sardinia-test.R
updated `soure()` of default ggplot theme

@@ -13,7 +13,7 @@ library('gratia')
 source('functions/betals.r') # custom beta location-scale family
 source('functions/scale-ndvi.R')
 source('functions/ndvi-palette.R')
-source('analysis/figures/default-ggplot-theme.R')
+source('analysis/figures/000-default-ggplot-theme.R')
 theme_set(theme_bw())
 
 # the bounding box for sardinia (large enough to include all coast)

### 2025-04-01 analysis/003-merging-rasters.R

@@ -198,7 +198,7 @@ tictoc::toc()
 # check the first raster ----
 if(FALSE) {
   library('ggplot2')
-  source('analysis/figures/default-ggplot-theme.R')
+  source('analysis/figures/000-default-ggplot-theme.R')
   
   dates_aggr %>%
     filter(group == group[1]) %>%

### 2025-04-01 analysis/010-test-hgams.R

@@ -6,7 +6,7 @@ library('sf')      # for simple fleature objects
 library('terra')   # for rasters
 library('elevatr') # for extracting elevation
 library('gratia')  # for plotting GAMs
-source('analysis/figures/default-ggplot-theme.R')
+source('analysis/figures/000-default-ggplot-theme.R')
 source('functions/add_nb.R')
 
 ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')

### 2025-04-01 analysis/figures/010-input-data.R

@@ -4,7 +4,7 @@ library('terra')     # for rasters
 library('ggplot2')   # for fancy plots
 library('tidyterra') # for plotting spatRasters with ggplot2
 library('khroma')    # for colorblind-friendly palettes
-source('analysis/figures/default-ggplot-theme.R')
+source('analysis/figures/000-default-ggplot-theme.R')
 
 eco <- read_sf('data/ecoregions/ecoregions-polygons.shp') %>%
   filter(WWF_MHTNAM != 'Inland Water') %>%

### 2025-04-02 analysis/000-sardinia-test.R
added machine-dependent n of cores

@@ -38,9 +38,9 @@ hist(sardinia$area_km2)
 if(file.exists('data/sardinia-test/sardinia-ndvi.rds')) {
   sardinia_ndvi <- readRDS('data/sardinia-test/sardinia-ndvi.rds')
 } else {
-  if(.Platform$OS.type != 'unix') stop('NOAA rasters are on the lab Linux.')
+  if(.Platform$OS.type != 'unix') stop('NOAA rasters are on the H: Drive, and you may want to use more cores.')
   future::availableCores(logical = FALSE)
-  plan(multisession, workers = 30)
+  plan(multisession, workers = min(30, availableCores(logical = FALSE) - 2))
   sardinia_ndvi <-
     list.files(path = '/home/shared/NOAA_Files/',
                pattern = '.nc',
@@ -108,9 +108,11 @@ if(file.exists('models/sardinia-test/gaussian-gam.rds')) {
 draw(m_gaus, rug = FALSE)
 ggsave('figures/sardinia-test/sardinia-ndvi-gaussian-terms.png',
        width = 9, height = 6, units = 'in', dpi = 300, bg = 'white')
+
 draw(m_gaus, rug = FALSE, fun = \(x) ndvi_to_11(x + coef(m_gaus)['(Intercept)']))
 ggsave('figures/sardinia-test/sardinia-ndvi-gaussian-terms-ndvi-scale.png',
        width = 9, height = 6, units = 'in', dpi = 300, bg = 'white')
+
 summary(m_gaus)
 
 # fit a spatially explicit test model with a beta family ----
@@ -134,11 +136,13 @@ if(file.exists('models/sardinia-test/beta-gam.rds')) {
 draw(m_beta, rug = FALSE)
 ggsave('figures/sardinia-test/sardinia-ndvi-beta-terms.png',
        width = 9, height = 6, units = 'in', dpi = 300, bg = 'white')
+
 draw(m_beta, rug = FALSE,
      fun = \(x) m_beta$family$linkinv(x + coef(m_beta)['(Intercept)']) %>%
        ndvi_to_11())
 ggsave('figures/sardinia-test/sardinia-ndvi-beta-terms-ndvi-scale.png',
        width = 9, height = 6, units = 'in', dpi = 300, bg = 'white')
+
 summary(m_beta)
 
 # fit a spatially explicit test model with a betals family ----

### 2025-04-02 analysis/011-mean-ndvi-hgam.R
wrote code for mean ndvi hgam

@@ -1,0 +1,129 @@
+library('dplyr')   # for data wrangling
+library('sf')      # for shapefiles
+library('mgcv')    # for Genralized Additive Models
+library('sf')      # for simple fleature objects
+library('terra')   # for rasters
+library('gratia')  # for plotting GAMs
+source('analysis/figures/000-default-ggplot-theme.R')
+source('functions/add_nb.R')
+
+ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
+
+# resolution is (0.05 degrees)^2 = (3 degree minutes)^2
+# create raster of ecoregion
+r_eco <- rast('data/ecoregions/wwf-ecoregions.tif')
+
+# create raster of polygon ID
+r_poly_id <- rast('data/ecoregions/ecoregion-polygon-id.tif')
+
+# get raster of elevations
+r_elev <- rast('data/elev-raster.tif')
+
+# create raster of distance from coast
+r_dist <- rast('data/distance-from-coast-m.tif')
+
+# set up the data for the model
+if(file.exists('data/hbam-ndvi-data.rds')) {
+  d <- readRDS('data/hbam-ndvi-data.rds')
+} else {
+  d <- readRDS('data/ndvi-global-15-day-average.rds') %>%
+    #' use `exact_extract()` for more data?
+    mutate(
+      doy = lubridate::yday(central_date),
+      poly_id = exactextract::exact_extract(r_poly_id, data.frame(x, y))[, 2], # 1 is pixel ID
+      wwf_ecoregion = exactextract::exact_extract(r_eco, data.frame(x, y))[, 2],
+      # distance_coast_m = exactextract::exact_extract(r_dist, data.frame(x, y))[, 2],
+      elevation_m = exactextract::exact_extract(r_elev, data.frame(x, y))[, 2]) %>%
+    # convert strings to factors
+    mutate(wwf_ecoregion = paste(wwf_ecoregion, if_else(y < 0, 'S', 'N')),
+           wwf_ecoregion = factor(wwf_ecoregion))
+  
+  apply(d, \(.row) any(is.na(.row))) %>%
+    mean()
+  
+  d <- na.omit(d)
+  
+  # set excessively low elevations to sea level (occurs mostly near islands)
[…]

### 2025-04-02 analysis/010-test-hgams.R
fixed file names, added comments of dev expl

@@ -141,8 +141,8 @@ d <- mutate(d, poly_id = factor(poly_id, levels = names(nbs)))
 # global smooths only
 # 2020 only: initial is 34 s, fit is 5 s
 # 2020 and 2021: initial is 40 s, fit is 10 s
-if(file.exists('models/global-test/m0-2020-2021.rds')) {
-  m0 <- readRDS('models/global-test/m0-2020-2021.rds')
+if(file.exists('models/global-test/m0-2020-2021-gam.rds')) {
+  m0 <- readRDS('models/global-test/m0-2020-2021-gam.rds')
 } else {
   m0 <- bam(ndvi_15_day_mean ~
               s(elevation_m, bs = 'cr', k = 5) +
@@ -155,7 +155,7 @@ if(file.exists('models/global-test/m0-2020-2021.rds')) {
             discrete = TRUE,
             control = gam.control(nthreads = 1, trace = TRUE))
   draw(m0, rug = FALSE)
-  saveRDS(m0, 'models/global-test/m0-2020-2021.rds')
+  saveRDS(m0, 'models/global-test/m0-2020-2021-gam.rds')
 }
 
 # adding s(doy, ecoregion); no common doy effect across all ecoregions
@@ -219,10 +219,11 @@ d %>%
 
 # testing ecoregion and polygon as fixed effects to hopefully decrease fitting time
 #' 2020 AND 2021: setup is 70 s, fit is 18 minutes
-if(file.exists('models/global-test/m3-2020-2021.rds')) {
-  m3 <- readRDS('models/global-test/m3-2020-2021.rds')
+if(file.exists('models/global-test/m3-2020-2021-gam.rds')) {
+  m3 <- readRDS('models/global-test/m3-2020-2021-gam.rds')
 } else {
   m3 <- bam(ndvi_15_day_mean ~
+              wwf_ecoregion +
               s(elevation_m, bs = 'cr', k = 5) +
               s(log(distance_coast_m), bs ='ad', k = 10) +
               s(doy, by = wwf_ecoregion, bs = 'cc', k = 10) +
@@ -234,11 +235,20 @@ if(file.exists('models/global-test/m3-2020-2021.rds')) {
             discrete = TRUE,
             control = gam.control(nthreads = 1, trace = TRUE))
   draw(m3, rug = FALSE)
-  saveRDS(m3, 'models/global-test/m3-2020-2021.rds')
[…]

### 2025-04-02 writing/manuscript.Rmd
updated after Ellen Aikens's talk

@@
 ---
-title: \large A new global measure of environmental unpredictability 
+title: \large **ACRONYM**: A new global measure of environmental unpredictability 
 author:
@@
 
+# Acronym:
+
+* NVGI: NDVI Variance GAM-smoothed Index
+* GAM-Estimated 
+* Ginli: GAM-Inferred NDVI ??? Index
+* env var
+
+potential words starting with a vowel: Index, Estimate(d), unpredictability
+
+potential words starting with a consonant: GAM, smooth(ed), NDVI, stochasticity, variance,
+
+Global GAM
+
 # Introduction
@@
 
-hex plots of mean and var vs common metrics (e.g., spp diversity, npp, hfi, ...)
+hex plots of mean and var vs common metrics:
+* temperature
+* seasonal temperature range
+* precip
+* seasonal temperature range
+* spp diversity
+* net primary productivity
+* hfi
+* max animal weight
 
@@
 
+* stable states
+* green wave surfing
+* migration routes are key habitat (Ortega et al., in prep), especially in stochastic environments
+* @sawyer_migratory_2019
+* https://www.sciencedirect.com/science/article/pii/S0960982220308484
+* https://onlinelibrary.wiley.com/doi/full/10.1111/gcb.15169
+* industrial development affects animals' ability to respond to both predictable and unpredictable change (https://www.nature.com/articles/s41559-022-01887-9)
+* effects of sudden deep snow in red desert on proghorn (https://www.sciencedirect.com/science/article/abs/pii/S0960982225002957)
+* https://www.nature.com/articles/s41467-023-37750-z
+* https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.4238
+
+* smoothness of mu impacts smoothness and size of s2: depends on animals' ability to respond to, learn, and predict E(R)
+
 # Useful references
 
-* `{khroma}`: doi:10.5281/zenodo.1472077
+* `{khroma}`: @frerebeau_khroma_2024
 * @keith_predicting_2008

### 2025-04-04 writing/appendix-1.Rmd
dropped dist figure, added color palettes fig

@@
 title: \large A new global measure of environmental unpredictability 
-subtitle: Appendix 1
+subtitle: Appendix A
 author:
@@
 header-includes:
+    - \renewcommand{\figurename}{Figure B\!\!} # for "Figure Bx."
     - \usepackage{setspace}\doublespacing # for double-spaced text
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, out.width = '100%')
+knitr::opts_chunk$set(echo = FALSE, out.width = '100%', fig.align = 'center')
 ```
@@
 The code for generating the figures is available [on GitHub](https://github.com/QuantitativeEcologyLab/ndvi-stochasticity/blob/main/analysis/figures/input-data.R).
+The figure below illustrates each color palette for each type of vision (trichromatic, deuteranope, protanope, tritanope, and achomatic).
+Palette A is commonly attributed to Masataka Okabe and Kei Ito [@ichihara_color_2008], while the other palettes were created by Fabio Crameri [@crameri_geodynamic_2018; @crameri_scientific_2018].
+
+```{r color-palettes, fig.cap="Color palettes used in the figures below."}
+include_graphics('../figures/input-data/color-palettes.png')
+```
 
@@
 
-```{r polygons, fig.cap="Map of the polygons for each ecoregion. Each polygon is colored with one of 9 colors based on the polygon names. Note that some neighboring polygons have the same color."}
+```{r polygons, fig.cap="Map of the polygons for each ecoregion. Each polygon is colored randomly with one of the 9 colors. Note that some neighboring polygons have the same color."}
 include_graphics('../figures/input-data/polygons.png')
@@
 
-```{r distance-coast, fig.cap="Map of the estimated distance from the nearest coastal edge."}
-include_graphics('../figures/input-data/dist-m.png')
+```{r distance-coast, fig.cap="Map of the estimated distance from the nearest coastal edge.", eval=FALSE}
+include_graphics('../figures/input-data/dist-km.png')
 ```

### 2025-04-04 writing/manuscript.Rmd
dropped dist figure, added color palettes fig

@@
 
-maps of input data
+## Choice of color schemes
 
-NDVI palette is colorblid-friendly
+We represent NDVI using a modified version of Fabio Crameri's divergent *bukavu* palette [@crameri_ref], which has high-contrast for deuteranope and protanope vision.
+For NDVI values between 0 and 1 (the right half of the palette), the colors have sufficient contrast for both tritanope and achomatic vision.
+Appendix A contains an approximate representation of the color palette for each vision type.
+
+## Input data
+
+Maps are in Aappendix A.
 
@@
 
-* stable states
+* detecting state shifts
 * green wave surfing
@@
 * https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.4238
+* see data for snow and pronghorn paper (Aikens et al. 2025)
 

### 2025-04-04 analysis/figures/010-input-data.R

@@ -2,6 +2,7 @@ library('dplyr')     # for data wrangling
 library('sf')        # for simple features
 library('terra')     # for rasters
 library('ggplot2')   # for fancy plots
+library('cowplot')   # for fancy plots in grids
 library('tidyterra') # for plotting spatRasters with ggplot2
 library('khroma')    # for colorblind-friendly palettes
 source('analysis/figures/000-default-ggplot-theme.R')
@@ -73,15 +74,19 @@ if(FALSE) {
 n_distinct(eco$poly_id)
 pal_polys <- c(color('okabeito')(8), 'grey')
 plot_scheme_colorblind(pal_polys)
+ggsave('polygons-palette.png', path = 'figures/input-data',
+       width = 6, height = 4, units = 'in', dpi = 300, bg = 'white')
 
+set.seed(-3)
 p_poly <-
   ggplot(eco) +
-  geom_sf(fill = pal_polys[1:nrow(eco) %% length(pal_polys) + 1],
-          lwd = .05) +
+  geom_sf(fill = sample(pal_polys, nrow(eco), replace = TRUE),
+          lwd = .15, color = 'white') +
+  geom_sf(fill = 'transparent', lwd = .05, color = 'black') +
   scale_x_continuous(expand = c(0, 0))
 
 ggsave('figures/input-data/polygons.png', p_poly,
-       width = 10, height = 4.4, units = 'in', dpi = 600, bg = 'white')
+       width = 10, height = 4.4, units = 'in', dpi = 1200, bg = 'white')
 
 # ecoregions ----
 n_distinct(eco$WWF_MHTNAM)
@@ -164,21 +169,35 @@ p_n
 ggsave('figures/input-data/n-rasters-time.png', p_n,
        width = 8, height = 5, units = 'in', dpi = 300, bg = 'white')
 
-# north/south hemispheres
-ns <- rast('data/') %>%
-  as.data.frame(xy = TRUE) %>%
-  transmute(x, y, z = if_else(y < 0, 'S', 'N'))
+#' function similar to `khroma::plot_scheme_colorblind()`
+plot_pal <- function(pal) {
+  .d <- tibble(x = floor((1:length(pal))),
+               y = x %% 2)
+  
+  .p <- colorblindr::palette_plot(pal, color_labels = FALSE,
+                                  xmargin = 0)
+  
+  plot_grid(.p +
+              ggtitle('Trichromatic') +
[…]

### 2025-04-04 writing/appendix-1.Rmd
re-shuffled figures and added Okabe & Ito ref

@@
 The figure below illustrates each color palette for each type of vision (trichromatic, deuteranope, protanope, tritanope, and achomatic).
-Palette A is commonly attributed to Masataka Okabe and Kei Ito [@ichihara_color_2008], while the other palettes were created by Fabio Crameri [@crameri_geodynamic_2018; @crameri_scientific_2018].
+Palette A was created by Masataka Okabe and Kei Ito [@okabe_color_2008], while the other palettes were created by Fabio Crameri [@crameri_geodynamic_2018; @crameri_scientific_2018].
 
@@
 
+```{r polygons, fig.cap="Map of the polygons for each ecoregion. Each polygon is colored randomly with one of the 9 colors. Note that some neighboring polygons have the same color."}
+include_graphics('../figures/input-data/polygons.png')
+```
+
 ```{r ecoregions, fig.cap="Map of the ecoregions."}
 include_graphics('../figures/input-data/ecoregions.png')
-```
-
-```{r polygons, fig.cap="Map of the polygons for each ecoregion. Each polygon is colored randomly with one of the 9 colors. Note that some neighboring polygons have the same color."}
-include_graphics('../figures/input-data/polygons.png')
 ```

### 2025-04-11 analysis/000-sardinia-test.R
remove unnecessary `theme_bw()`

@@ -14,7 +14,6 @@ source('functions/betals.r') # custom beta location-scale family
 source('functions/scale-ndvi.R')
 source('functions/ndvi-palette.R')
 source('analysis/figures/000-default-ggplot-theme.R')
-theme_set(theme_bw())
 
 # the bounding box for sardinia (large enough to include all coast)
 sardinia_bbox <- tibble(x = c(7.6, 10.4), y = c(38.8, 41.3)) %>%

### 2025-04-11 analysis/misc/test-openblas.R
rename `analysis/misc/test-openMP.R`

@@ -1,4 +1,3 @@
-#' *could not set up parallelization correctly*
 # https://stat.ethz.ch/R-manual/R-devel/library/mgcv/html/mgcv-parallel.html
 
 ## illustration of multi-threading with gam...
@@ -8,11 +7,11 @@ dat <- gamSim(1,n=1e6,dist="poisson",scale=.1)
 k <- 12;bs <- "cr";ctrl <- gam.control(nthreads=10, trace = TRUE)
 
 times <- rep(NA_real_, 5)
-names(times) <- c('gam', 'gam w 2 threads', 'bam', 'bam w 2 threads in control',
-                  'bam w 2 threads in nthreads')
+names(times) <- c('gam', 'gam w 10 threads', 'bam', 'bam w 10 threads in control',
+                  'bam w 10 threads in nthreads')
 
 times[1] <- system.time(b1<-gam(y~s(x0,bs=bs)+s(x1,bs=bs)+s(x2,bs=bs,k=k),
-                                family=poisson,data=dat,method="REML"))
+                                family=poisson,data=dat,method="REML"))[3]
 
 times[2] <- system.time(b2<-gam(y~s(x0,bs=bs)+s(x1,bs=bs)+s(x2,bs=bs,k=k),
                                 family=poisson,data=dat,method="REML",control=ctrl))[3]
@@ -29,10 +28,10 @@ times[5] <- system.time(b5<-bam(y~s(x0,bs=bs)+s(x1,bs=bs)+s(x2,bs=bs,k=k),
 # repeat with a larger dataset
 library(parallel)
 set.seed(9)
-dat_large <- gamSim(1,n=1e7,dist="poisson",scale=.1)
+dat_large <- gamSim(1,n=1e8,dist="poisson",scale=.1)
 
 times_large <- rep(NA_real_, 3)
-names(times_large) <- c('bam', 'bam with cluster', 'bam w 2 threads in nthreads')
+names(times_large) <- c('bam', 'bam with cluster', 'bam w 10 threads in nthreads')
 times_large
 
 cl <- makeCluster(5) # 5-core cluster
@@ -42,11 +41,10 @@ times_large[1] <- system.time(b3<-bam(y~s(x0,bs=bs)+s(x1,bs=bs)+s(x2,bs=bs,k=k),
                                       discrete=TRUE))[3]
 times_large[2] <- system.time(b4<-bam(y~s(x0,bs=bs)+s(x1,bs=bs)+s(x2,bs=bs,k=k),
[…]

### 2025-04-11 analysis/011-mean-ndvi-hgam.R
making `elev_m > 0`; saving `nbs` object

@@ -43,7 +43,9 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
   
   d <- na.omit(d)
   
-  # set excessively low elevations to sea level (occurs mostly near islands)
+  # some excessively low elevations for coastal pixels sea level
+  # occurs mostly near islands; leaving values as they are because it helps
+  # inform about the lower NDVI due to the coast
   if(FALSE) {
     plot(st_geometry(ecoregions))
     d %>%
@@ -51,7 +53,7 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
       points(y ~ x, ., col = 'red', pch = 19)
   }
   
-  d <- mutate(d, elevation_m = if_else(elevation_m < -20, 0, elevation_m))
+  d <- mutate(d, elevation_m = if_else(elevation_m < 0, 0, elevation_m))
   
   saveRDS(d, 'data/hbam-ndvi-data.rds')
 }
@@ -71,34 +73,40 @@ if(FALSE) {
 
 ecoregions <- filter(ecoregions, in_data | n_neigh > 0)
 
-nbs <-
-  spdep::poly2nb(pl = ecoregions,
-                 row.names = ecoregions$poly_id, # so that names match IDs
-                 queen = TRUE) # 1 point in common is sufficient
-length(nbs) == nrow(ecoregions)
-names(nbs) <- attr(nbs, 'region.id') # names do not need to match indices
-
-# add missing neighbors because of polygons that cross the edge of the map
-c('poly 8225', 'poly 8038', 'poly 7987', 'poly 8226', 'poly 7988', 'poly 8040') %in%
-  ecoregions$poly_id
-
-layout(matrix(1:6, ncol = 2))
-# 7987 and 8038 are already neighbors
-# 7988 and 8040 are already neighbors
-# 7987 and 8226 do not touch
-add_nb(p1 = 'poly 8225', p2 = 'poly 8226', add = TRUE)
-add_nb(p1 = 'poly 8038', p2 = 'poly 8040', add = TRUE)
-add_nb(p1 = 'poly 7987', p2 = 'poly 7988', add = TRUE)
-add_nb(p1 = 'poly 7987', p2 = 'poly 8040', add = TRUE)
-add_nb(p1 = 'poly 7988', p2 = 'poly 8038', add = TRUE)
[…]

### 2025-04-11 writing/appendix-A.Rmd
cleaned up appendix captions and figures

---
title: \large A new global measure of environmental unpredictability 
subtitle: Appendix A
author:
  - name: Stefano Mezzini
    institute: [braes, biol]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
    correspondence: "yes"
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'ndvi-stochasticity.bib'
csl: 'freshwater-biology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \renewcommand{\figurename}{Figure A\!\!} # for "Figure Ax."
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
    keep_tex: false # for submitting TeX file
editor_options:
  chunk_output_type: console
---

\clearpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '100%', fig.align = 'center')
```

```{r packages, warning=FALSE}
library('knitr')
```

\noindent Figure A1 below illustrates the color palettes used in this appendix, for each type of vision (trichromatic, deuteranope, protanope, tritanope, and achomatic). Palette A was created by Masataka Okabe and Kei Ito [@okabe_color_2008], while the other palettes were created by Fabio Crameri [@crameri_geodynamic_2018; @crameri_scientific_2018]. All palettes below are available via the `khroma` package [@frerebeau_khroma_2024] for `R` [@r_core_team_r_2024]. Figures A2-A5 show some of the data used for fitting the models for estimating mean NDVI and the variance around the mean. The code for generating the figures is available [on GitHub](https://github.com/QuantitativeEcologyLab/ndvi-stochasticity/blob/main/analysis/figures/input-data.R).

```{r color-palettes, fig.cap="Color palettes used in the figures below."}
include_graphics('../figures/input-data/color-palettes.png')
```

```{r polygons, fig.cap="Map of the polygons for each ecoregion. Each polygon is colored randomly with one of 9 colors. Note that some neighboring polygons have the same color."}
include_graphics('../figures/input-data/polygons.png')
```

```{r ecoregions, fig.cap="Map of the ecoregions. The Northern and Southern hemispheres were coded to have separate ecoregions in the model by appending \"N\" or \"S\" to the end of the ecoregion name."}
include_graphics('../figures/input-data/ecoregions.png')
```

```{r elevation, fig.cap="Map of elevation above sea level. All elevations below 0 m were set to 0 m to fix incorrect altitudes near coast lines. Terrestrial ecosystems with elevations below 0 m (such as the Dead Sea and the Qattara Depression) have also been set to 0 m because they have their own polygons, which allows to account for any differences in the Markov Random Field smooth."}
include_graphics('../figures/input-data/elev-m.png')
```

```{r distance-coast, fig.cap="Map of the estimated distance from the nearest coastal edge.", eval=FALSE}
# not evaluated!
include_graphics('../figures/input-data/dist-km.png')
```

```{r n-rasters, fig.cap="Number of days with a raster within 15-day periods starting on January $1^{\\text{st}}$, for each year. Cells with less than 15 rasters indicate one or more missing rasters for that 15-day period, which the exception of cells near day 366, which have 6 days during leap years ($366 \\mod 15 = 6$) and 5 days otherwise ($365 \\mod 15 = 5$)."}
include_graphics('../figures/input-data/n-rasters-time.png')
```

\clearpage

# References

### 2025-04-11 analysis/figures/010-input-data.R

@@ -106,11 +106,20 @@ ggsave('figures/input-data/ecoregions.png', p_eco,
 r_elev <- rast('data/elev-raster.tif') %>%
   crop(., st_transform(eco, crs(.)), mask = TRUE)
 
-hist(r_elev)
-values(r_elev) <- if_else(values(r_elev) < -400, -400, values(r_elev))
-
+# change coastal values to 0 m above sea level
 ggplot() +
   geom_spatraster(data = r_elev < 0)
+
+# qattara depression and dead sea are < 0 m but also separate polygons
+ggplot() +
+  geom_spatraster(data = r_elev < 0) +
+  scale_fill_manual(values = c('white', 'red'))
+
+hist(r_elev, breaks = 1000)
+abline(v = 0, col = 'red')
+mean(values(r_elev < 0), na.rm = TRUE)
+
+values(r_elev) <- if_else(values(r_elev) < 0, 0, values(r_elev))
 
 p_elev <-
   ggplot() +
@@ -167,7 +176,7 @@ p_n <-
 p_n
 
 ggsave('figures/input-data/n-rasters-time.png', p_n,
-       width = 8, height = 5, units = 'in', dpi = 300, bg = 'white')
+       width = 10, height = 5, units = 'in', dpi = 300, bg = 'white')
 
 #' function similar to `khroma::plot_scheme_colorblind()`
 plot_pal <- function(pal) {
@@ -196,8 +205,8 @@ plot_grid(NULL,
           # plot_pal(color('davos')(1e3)),
           # NULL,
           plot_pal(rev(color('lapaz')(1e3))),
-          label_x = 0, ncol = 4,  rel_widths = c(0.03, 1),
-          labels = c('A.', '', 'B.', '', 'C.', '', 'D.', '', 'E.'))
+          label_x = 0, ncol = 4,  rel_widths = c(0.075, 1),
+          labels = c('A2.', '', 'A3.', '', 'A4.', '', 'A5.'))
 
 ggsave('figures/input-data/color-palettes.png', width = 8, height = 6,
        scale = 2, units = 'in', dpi = 300, bg = 'white')

### 2025-04-11 analysis/010-test-hgams.R
fixed poly names; removed `dist_coast_m`

@@ -1,3 +1,5 @@
+#' *THE MODELS FIT IN THIS SCRIPT ARE FOR INITIAL TESTING ONLY*
+#' *DO NOT USE THESE MODELS TO PREDICT*
 library('dplyr')   # for data wrangling
 library('purrr')   # for functional programming
 library('sf')      # for shapefiles
@@ -12,7 +14,7 @@ source('functions/add_nb.R')
 ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
 
 # resolution is (0.05 degrees)^2 = (3 degree minutes)^2
-r_0 <- rast('data/avhrr-viirs-ndvi/raster-files/AVHRR-Land_v005_AVH13C1_NOAA-07_19810624_c20170610041337.nc',
+r_0 <- rast('H:/GitHub/ndvi-stochasticity/data/avhrr-viirs-ndvi/raster-files/AVHRR-Land_v005_AVH13C1_NOAA-07_19810624_c20170610041337.nc',
             lyr = 'QA') %>% # to have a value for each cell irrespective of raster
   aggregate(2) # because aggregated in the dataset
 values(r_0) <- 1
@@ -91,11 +93,11 @@ if(file.exists('data/hbam-ndvi-data-2020-2021-only.rds')) {
 if(FALSE) {
   plot(st_geometry(ecoregions))
   d %>%
-    filter(elevation_m < -20) %>%
+    filter(elevation_m < 0) %>%
     points(y ~ x, ., col = 'red', pch = 19)
 }
 
-d <- mutate(d, elevation_m = if_else(elevation_m < -20, 0, elevation_m))
+d <- mutate(d, elevation_m = if_else(elevation_m < 0, 0, elevation_m))
 
 # create list of neighbors ----
 ecoregions$in_data <- ecoregions$poly_id %in% unique(d$poly_id)
@@ -116,11 +118,13 @@ nbs <- spdep::poly2nb(pl = ecoregions,
                       row.names = ecoregions$poly_id, # so that names match IDs
                       queen = TRUE) # 1 point in common is sufficient
 length(nbs) == nrow(ecoregions)
-names(nbs) <- attr(nbs, 'region.id') # names do not need to match indices
+names(nbs) <- ecoregions$poly_id # names do not need to match indices
 
 # add missing neighbors because of polygons that cross the edge of the map
 c('poly 8225', 'poly 8038', 'poly 7987', 'poly 8226', 'poly 7988', 'poly 8040') %in%
   ecoregions$poly_id
[…]

### 2025-04-11 analysis/011-mean-ndvi-hgam.R
not using `exactextractr`; finalized mean model

@@ -9,39 +9,43 @@ source('functions/add_nb.R')
 
 ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
 
-# resolution is (0.05 degrees)^2 = (3 degree minutes)^2
-# create raster of ecoregion
-r_eco <- rast('data/ecoregions/wwf-ecoregions.tif')
-
-# create raster of polygon ID
-r_poly_id <- rast('data/ecoregions/ecoregion-polygon-id.tif')
-
-# get raster of elevations
-r_elev <- rast('data/elev-raster.tif')
-
-# create raster of distance from coast
-r_dist <- rast('data/distance-from-coast-m.tif')
-
 # set up the data for the model
 if(file.exists('data/hbam-ndvi-data.rds')) {
   d <- readRDS('data/hbam-ndvi-data.rds')
 } else {
-  d <- readRDS('data/ndvi-global-15-day-average.rds') %>%
-    #' use `exact_extract()` for more data?
-    mutate(
-      doy = lubridate::yday(central_date),
-      poly_id = exactextract::exact_extract(r_poly_id, data.frame(x, y))[, 2], # 1 is pixel ID
-      wwf_ecoregion = exactextract::exact_extract(r_eco, data.frame(x, y))[, 2],
-      # distance_coast_m = exactextract::exact_extract(r_dist, data.frame(x, y))[, 2],
-      elevation_m = exactextract::exact_extract(r_elev, data.frame(x, y))[, 2]) %>%
-    # convert strings to factors
-    mutate(wwf_ecoregion = paste(wwf_ecoregion, if_else(y < 0, 'S', 'N')),
-           wwf_ecoregion = factor(wwf_ecoregion))
+  # resolution is (0.05 degrees)^2 = (3 degree minutes)^2
+  # create raster of ecoregion
+  r_eco <- rast('data/ecoregions/wwf-ecoregions.tif')
   
-  apply(d, \(.row) any(is.na(.row))) %>%
-    mean()
+  # create raster of polygon ID
+  r_poly_id <- rast('data/ecoregions/ecoregion-polygon-id.tif')
+  
+  # get raster of elevations
+  r_elev <- rast('data/elev-raster.tif')
+  
+  # create raster of distance from coast
+  r_dist <- rast('data/distance-from-coast-m.tif')
+  
[…]

### 2025-04-11 analysis/011-mean-ndvi-hgam.R
changed all `m_fe` to `m`

@@ -124,7 +124,7 @@ SAMPLE <- sample(1:(nrow(d) / 3),      # only first part of the years
 if(file.exists('models/global-models/hbam-mean-ndvi-DATE.rds')) {
   m <- readRDS(paste0('models/global-models/hbam-mean-ndvi-DATE.rds'))
 } else {
-  m_fe <- bam(
+  m <- bam(
     ndvi_15_day_mean ~
       wwf_ecoregion + # to avoid intercept shrinkage
       s(poly_id, bs = 'mrf', xt = list(nb = nbs)) +
@@ -143,24 +143,24 @@ if(file.exists('models/global-models/hbam-mean-ndvi-DATE.rds')) {
     nthreads = 50,
     control = gam.control(trace = TRUE))
   
-  saveRDS(m_fe, paste0('models/global-models/mean-ndvi-hbam-', DATE, '.rds'))
+  saveRDS(m, paste0('models/global-models/mean-ndvi-hbam-', DATE, '.rds'))
 }
 
-#' plot each smooth separately
-p_hbam_doy <- draw(m_fe, rug = FALSE,
-                   select = which(grepl('s(doy', smooths(m_fe))))
+# plot each smooth separately
+p_hbam_doy <- draw(m, rug = FALSE,
+                   select = which(grepl('s(doy', smooths(m))))
 ggsave(paste0('figures/hbam-mean-ndvi-fe-doy-', DATE, '.png'),
        plot = p_hbam_doy, width = 24, height = 36, units = 'in', dpi = 300,
        bg = 'white')
 
-p_hbam_y <- draw(m_fe, rug = FALSE,
-                 select = which(grepl('s(year', smooths(m_fe))))
+p_hbam_y <- draw(m, rug = FALSE,
+                 select = which(grepl('s(year', smooths(m))))
 ggsave(paste0('figures/hbam-mean-ndvi-fe-year-', DATE, '.png'),
        plot = p_hbam_y, width = 24, height = 36, units = 'in', dpi = 300,
        bg = 'white')
 
-p_hbam_elev <- draw(m_fe, rug = FALSE,
-                    select = which(smooths(m_fe) == 's(elev_m)'))
+p_hbam_elev <- draw(m, rug = FALSE,
+                    select = which(smooths(m) == 's(elev_m)'))
 ggsave(paste0('figures/hbam-mean-ndvi-fe-elev_m-', DATE, '.png'),
        plot = p_hbam_elev, width = 4, height = 6, units = 'in', dpi = 300,
        bg = 'white')

### 2025-04-11 analysis/012-var-ndvi-hgam.R
wrote code for variance model

@@ -1,0 +1,70 @@
+library('dplyr')   # for data wrangling
+library('sf')      # for shapefiles
+library('mgcv')    # for Genralized Additive Models
+library('sf')      # for simple fleature objects
+library('terra')   # for rasters
+library('gratia')  # for plotting GAMs
+source('analysis/figures/000-default-ggplot-theme.R')
+
+ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
+
+# import the data for the model
+d <- readRDS('data/hbam-var-ndvi-data.rds')
+nbs <- readRDS('data/ecoregions/poly-nbs-global.rds')
+
+if(length(list.files('models/global-test', 'hbam-var-ndvi-*')) > 1) {
+  DATE <- '2025-04-XX'
+  m <- readRDS(paste0('models/global-test/hbam-var-ndvi-', DATE, '.rds'))
+} else {
+  m <- bam(
+    e_2 ~
+      wwf_ecoregion + # to avoid intercept shrinkage
+      s(poly_id, bs = 'mrf', xt = list(nb = nbs)) +
+      s(doy, wwf_ecoregion, bs = 'fs', xt = list(bs = 'cc'), k = 10) +
+      s(year, wwf_ecoregion, bs = 'fs', xt = list(bs = 'cr'), k = 10) +
+      ti(doy, year, wwf_ecoregion, bs = c('cc', 'cr', 're'), k = c(5, 5)) +
+      s(elevation_m, bs = 'cr', k = 5),
+    family = gaussian(),
+    data = d,
+    method = 'fREML',
+    knots = list(doy = c(0.5, 366.5)),
+    drop.unused.levels = TRUE, # FALSE slows down the model substantially
+    discrete = TRUE,
+    samfrac = 0.001, # find intial guesses with a subset of the data
+    nthreads = 50,
+    control = gam.control(trace = TRUE))
+  
+  saveRDS(m, paste0('models/global-test/hbam-var-ndvi-', DATE, '.rds'))
+}
+
+# plot each smooth separately
+p_hbam_doy <- draw(m, rug = FALSE,
+                   select = which(grepl('s(doy', smooths(m))))
+ggsave(paste0('figures/hbam-var-ndvi-fe-doy-', DATE, '.png'),
+       plot = p_hbam_doy, width = 24, height = 36, units = 'in', dpi = 300,
+       bg = 'white')
+
[…]

### 2025-04-11 analysis/010-test-hgams.R
added `dev.expl` for each model

@@ -260,7 +260,7 @@ tibble(model = 0:3,
        }))
 
 # model dev_expl terms
-#   m0      XXXX s(elevation_m), s(doy)
-#   m1      XXXX s(elevation_m), s(doy,wwf_ecoregion)
-#   m2      XXXX s(elevation_m), s(doy,wwf_ecoregion), s(poly_id)
-#   m3      XXXX s(elevation_m), s(doy):wwf_ecoregion + wwf_ecoregion + poly_id
+#   m0      14.3 s(elevation_m), s(doy)
+#   m1      57.5 s(elevation_m), s(doy,wwf_ecoregion)
+#   m2      67.9 s(elevation_m), s(doy,wwf_ecoregion), s(poly_id)
+#   m3      67.9 s(elevation_m), s(doy):wwf_ecoregion + wwf_ecoregion + poly_id

### 2025-04-11 analysis/011-mean-ndvi-hgam.R
removed test code

@@ -117,10 +117,6 @@ d <- mutate(d, poly_id = factor(poly_id, levels = names(nbs)))
 
 DATE <- format(Sys.Date(), '%Y-%m-%d')
 
-#' *TEMPORARY VARIABLE FOR PROPORTION OF DATA*
-SAMPLE <- sample(1:(nrow(d) / 3),      # only first part of the years
-                 floor(nrow(d) / 1e4)) # only a subset of the rows
-
 if(file.exists('models/global-models/hbam-mean-ndvi-DATE.rds')) {
   m <- readRDS(paste0('models/global-models/hbam-mean-ndvi-DATE.rds'))
 } else {
@@ -133,8 +129,7 @@ if(file.exists('models/global-models/hbam-mean-ndvi-DATE.rds')) {
       ti(doy, year, wwf_ecoregion, bs = c('cc', 'cr', 're'), k = c(5, 5)) +
       s(elevation_m, bs = 'cr', k = 5),
     family = gaussian(),
-    #' *TEMPORARY TEST (\/ DELETE)*
-    data = d[1:(nrow(d) / 3), ],
+    data = d,
     method = 'fREML',
     knots = list(doy = c(0.5, 366.5)),
     drop.unused.levels = TRUE,

### 2025-04-11 analysis/011-mean-ndvi-hgam.R
using all but 2 cores to fit the models

@@ -135,7 +135,7 @@ if(file.exists('models/global-models/hbam-mean-ndvi-DATE.rds')) {
     drop.unused.levels = TRUE,
     discrete = TRUE,
     samfrac = 0.001, # find intial guesses with a subset of the data
-    nthreads = 50,
+    nthreads = future::availableCores(logical = FALSE) - 2,
     control = gam.control(trace = TRUE))
   
   saveRDS(m, paste0('models/global-models/mean-ndvi-hbam-', DATE, '.rds'))

### 2025-04-11 analysis/012-var-ndvi-hgam.R

@@ -31,7 +31,7 @@ if(length(list.files('models/global-test', 'hbam-var-ndvi-*')) > 1) {
     drop.unused.levels = TRUE, # FALSE slows down the model substantially
     discrete = TRUE,
     samfrac = 0.001, # find intial guesses with a subset of the data
-    nthreads = 50,
+    nthreads = future::availableCores(logical = FALSE) - 2,
     control = gam.control(trace = TRUE))
   
   saveRDS(m, paste0('models/global-test/hbam-var-ndvi-', DATE, '.rds'))

### 2025-04-11 analysis/010-test-hgams.R
moved raster creation to data creation `if()`

@@ -13,68 +13,68 @@ source('functions/add_nb.R')
 
 ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
 
-# resolution is (0.05 degrees)^2 = (3 degree minutes)^2
-r_0 <- rast('H:/GitHub/ndvi-stochasticity/data/avhrr-viirs-ndvi/raster-files/AVHRR-Land_v005_AVH13C1_NOAA-07_19810624_c20170610041337.nc',
-            lyr = 'QA') %>% # to have a value for each cell irrespective of raster
-  aggregate(2) # because aggregated in the dataset
-values(r_0) <- 1
-plot(r_0)
-res(r_0)
-
-# create raster of ecoregion
-if(file.exists('data/ecoregions/wwf-ecoregions.tif')) {
-  r_eco <- rast('data/ecoregions/wwf-ecoregions.tif')
-} else {
-  r_eco <-
-    ecoregions %>%
-    select(WWF_MHTNAM) %>%
-    vect() %>%
-    rasterize(r_0, field = 'WWF_MHTNAM')
-  plot(r_eco, legend = FALSE)
-  writeRaster(r_eco, 'data/ecoregions/wwf-ecoregions.tif')
-}
-
-# create raster of polygon ID
-if(file.exists('data/ecoregions/ecoregion-polygon-id.tif')) {
-  r_poly_id <- rast('data/ecoregions/ecoregion-polygon-id.tif')
-} else {
-  r_poly_id <-
-    ecoregions %>%
-    select(poly_id) %>%
-    vect() %>%
-    rasterize(r_0, field = 'poly_id')
-  plot(r_poly_id, legend = FALSE)
-  writeRaster(r_poly_id, 'data/ecoregions/ecoregion-polygon-id.tif')
-}
-
-# get raster of elevations
-if(file.exists('data/elev-raster.tif')) {
-  r_elev <- rast('data/elev-raster.tif')
-} else {
-  r_elev <- get_elev_raster(ecoregions, z = 3) #'`z=3` gives `0.076 < res(r_0)`
-  r_elev <- rast(r_elev) #' `{elevatr}` v.0.99.0 still uses `raster::raster()`
-  res(r_elev)
-  res(r_0)
-  plot(r_elev)
-  r_elev <- project(r_elev, r_0) # project to common projection and resolution
-  writeRaster(r_elev, 'data/elev-raster.tif')
-}
-
-# create raster of distance from coast
[…]

### 2025-04-11 analysis/011-mean-ndvi-hgam.R
fixed creation of `nbs` (names)

@@ -90,7 +90,7 @@ if(file.exists('data/ecoregions/poly-nbs-global.rds')) {
                    row.names = ecoregions$poly_id, # so that names match IDs
                    queen = TRUE) # 1 point in common is sufficient
   length(nbs) == nrow(ecoregions)
-  names(nbs) <- attr(nbs, 'region.id') # names do not need to match indices
+  names(nbs) <- ecoregions$poly_id # names do not need to match indices
   
   # add missing neighbors because of polygons that cross the edge of the map
   c('poly 8225', 'poly 8038', 'poly 7987', 'poly 8226', 'poly 7988', 'poly 8040') %in%
@@ -111,7 +111,7 @@ if(file.exists('data/ecoregions/poly-nbs-global.rds')) {
 }
 
 # ensure names match the factor levels
-all(attr(nbs, 'region.id') == ecoregions$poly_id)
+all(names(nbs) == levels(d$poly_id))
 
 d <- mutate(d, poly_id = factor(poly_id, levels = names(nbs)))
 

### 2025-04-11 analysis/010-test-hgams.R
fitting test variance model

@@ -260,7 +260,46 @@ tibble(model = 0:3,
        }))
 
 # model dev_expl terms
-#   m0      14.3 s(elevation_m), s(doy)
-#   m1      57.5 s(elevation_m), s(doy,wwf_ecoregion)
-#   m2      67.9 s(elevation_m), s(doy,wwf_ecoregion), s(poly_id)
-#   m3      67.9 s(elevation_m), s(doy):wwf_ecoregion + wwf_ecoregion + poly_id
+#   m0    14.3 % s(elevation_m), s(doy)
+#   m1    57.5 % s(elevation_m), s(doy,wwf_ecoregion)
+#   m2    67.9 % s(elevation_m), s(doy,wwf_ecoregion), s(poly_id)
+#   m3    67.9 % s(elevation_m), s(doy):wwf_ecoregion + wwf_ecoregion + poly_id
+
+# test for variance -----
+# add fitted values and residuals to the dataset
+d$mu_hat <- fitted(m2)
+d$e <- residuals(m2)
+
+# check pixel-level mean residuals
+d <- d %>%
+  group_by(x, y) %>%
+  mutate(mean_e = mean(e, na.rm = TRUE)) %>%
+  ungroup()
+
+d %>%
+  group_by(x, y) %>%
+  slice(1) %>%
+  pull(mean_e) %>%
+  hist(breaks = 50, main = 'Mean pixel-level residual')
+
+#' *IF REMOVING PIXEL-LEVEL MEAN RESIDUAL*
+d <- mutate(d,
+            e = e - mean_e,
+            e_2 = e^2, # squared residuals (mean e^2 is var)
+            mu_hat = mu_hat + mean_e)
+
+# fit the model for the variance
+m_var <- bam(ndvi_15_day_mean ~
+            s(elevation_m, bs = 'cr', k = 5) +
+            s(doy, wwf_ecoregion, bs = 'fs', xt = list(bs = 'cc'), k = 10) +
+            s(poly_id, bs = 'mrf', xt = list(nb = nbs)),
+          family = gaussian(),
+          data = d,
+          method = 'fREML',
+          knots = list(doy = c(0.5, 366.5)),
+          drop.unused.levels = FALSE,
+          discrete = TRUE,
+          nthreads = 10,
+          control = gam.control(trace = TRUE))
+draw(m_var, rug = FALSE, select = 1:2)
+saveRDS(m_var, 'models/global-test/m-var-2020-2021-gam.rds')

### 2025-04-11 writing/manuscript.Rmd
wrote abstract; added some structure to the intro

--- 2025-03-31 writing/manuscript.Rmd
@@
 ---
-title: \large A new global measure of environmental unpredictability 
+title: "DENVar: A global Dynamic Estimate of NDVI Variance"
 author:
@@
 ```{r, echo=FALSE}
-abstract <- ""
+# need to write abstract before cover page to state number of words in it
+abstract <- "Since the 1980s, remote-sensing estimates of habitat productivity such as the Normalized Differnece Vegetation Index (NDVI) have helped develop and support a variety of hypotheses and inform policy and industriy-related decisions.
+The free availability of such data has allowed widespread access and use of well-established metrics of habitat productivity.
+However, historically, users tended to focus solely on (predictable) trends in productivity.
+In recent years, people have been recognizing the importance of envrironmental unpredictability (stochasticity), particularly as climate change and human-induced rapid environmental change transform the ecosystems species have evolved in and adapted to.
+We adress this gap by presenting a new global Dynamic Estimate of NDVI Variance, *DENVar*.
+We estimate spatiotemporal trends in mean NDVI and the variance around the mean using Hierarchical Generalized Additive Models, which provide a flexible yet transparent model structure, unlike machine learning approaches such as neural netwoks.
+We show that DENVar can be used as a reliable proxy of environmental stochasticity, strongly correlates with ***XXX***, and can be used to test many hepotheses related to environmental stochasitcity, such as forage unpredictability, regime shifts, ...
+.
+We conclude by offering some considerations around quantifying and interpreting envirnomental stochasticity, and we provide a publicly and freely available shiny app for estimating mean NDVI and the variance around it, DENVar, for given coordinates and optional dates.
+The app requires no knowledge of statistics or coding.
+Additionally, we provide a temporally static raster of mean NDVI and DENVar for the years 1981-2025 that can be used for GIS applications and does not require any knowledge of $\\texttt{R}$."
+
+if(FALSE) {
+  stringi::stri_count_words(abstract)
+}
 ```
@@
 
-https://www.sciencedirect.com/science/article/abs/pii/S0034425718305625
-https://silvis.forest.wisc.edu/data/dhis/
+<!-- remote sensing allows us to study landscapes in detail and over large spatiotemporal scales -->
+
+\noindent Developments in remote sensing since the 1970s have greatly expanded our ability to quantify and monitor lanscapes using satellite imagery
+
+- NDVI is an estimate of greenness, NPP, forage availability
+  - NDVI has many uses and is often convenient because it's free, well-established, and
+
+<!-- successes of NDVI: pettorelli, Aikens -->
+
+many people focus on changes in the mean
+
+<!-- what counts as a change? predictable vs stochastic changes -->
+
+ability to predict and respond to changes in mean conditions (forage availability, phenology, etc.) depends on the scale of changes relative to the perception scale [@levin_problem_1992; @frankham_importance_2004; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020] and memory [@abrahms_memory_2019].
+Consequently, environmental stochasticity is also an important driver, besides changes in the mean [@mezzini_how_2025]
+
+<!-- difficulties in estimating variance -->
+
+* variance is hard to calculate, as it depends on a correct estimate of the mean
+* paper on estimating the asymptote of environmental variance globally, but variance for a closed system must converge towards a finite value [@bachmaier_variogram_2011; @fleming_fine-scale_2014]
+
+<!-- we present a new measure that is estimated in an understandable way -->
+
+* we present DENVar
+* HGAMs are higly flexible models that allow us to estimate trends in the data without imposing preconceived expectations
+* more transparent than black-box machine learning methods like neural networks, etc.
+* https://www.sciencedirect.com/science/article/abs/pii/S0034425718305625
+* https://silvis.forest.wisc.edu/data/dhis/
 
@@
 
-maps of input data
+## Choice of color schemes
 
-NDVI palette is colorblid-friendly
+We represent NDVI using a modified version of Fabio Crameri's divergent *bukavu* palette [@crameri_geodynamic_2018], which has high-contrast for deuteranope and protanope vision.
+For NDVI values between 0 and 1 (the right half of the palette), the colors also have sufficient contrast for the colors to be distinguishable by both tritanope and achomatic vision.
+Appendix A contains an approximate representation of the color palette for each vision type.
+We obtained all color palettes from the `khroma` package [v.
+1.14.0, @frerebeau_khroma_2024] for `R` [v.
+4.4.1, @r_core_team_r_2024].
+
+## Input data
+
+The models we fit
+
+### NDVI data
+
+* daily AVHRR-VIIRS
+* 15-day average and `aggregate(, 2)` to make dataset small enough to fit in a single data frame
+
+```{r, eval=FALSE}
+# approximate estimate of the number of unique DOY values
+length(unique(seq(1, 365.25 * (2025 - 1981), by = 15) %% c(365, 365, 365, 366)))
+```
+
+### Predictor data
+
+Maps are in Appendix A.
+
+## Modeling
+
+We estimated spatiotemporal trends in mean NDVI using Hierarchical Generalized Additive Models (HGAMs) via the `mgcv` package for `R` [v.
+1.9-3, @wood_generalized_2017].
+To reduce modeling fitting times with negligible losses to model accuracy, we used the `bam()` function with fast REstricted Marginal Likelihood (`method = fREML`) and covariate discretization (`discrete = TRUE`).
+The model structure is available below.
+
+```{r, eval=FALSE}
+# the model for the mean
+```
 
@@
 
-hex plots of mean and var vs common metrics (e.g., spp diversity, npp, hfi, ...)
+hex plots of mean and var vs common metrics:
+* temperature
+* seasonal temperature range
+* precip
+* seasonal temperature range
+* spp diversity
+* gross primary productivity
+* hfi
+* max animal weight
+* https://silvis.forest.wisc.edu/globalwui/
+* https://silvis.forest.wisc.edu/data/dhis/:
+  * NDVI16 (Normalized Difference Vegetation Index)
+  * EVI16 (Enhanced Vegetation Index)
+  * FPAR8 (Fraction absorbed Photosynthetically Active Radiation)
+  * LAI8 (Leaf Area Index)
+  * GPP8 (Gross Primary Productivity)
+* https://silvis.forest.wisc.edu/data/eurasia-trends/
 
@@
 
+* green wave surfing
+* migration routes are key habitat (Ortega et al., in prep), especially in stochastic environments
+* @sawyer_migratory_2019
+* https://www.sciencedirect.com/science/article/pii/S0960982220308484
+* https://onlinelibrary.wiley.com/doi/full/10.1111/gcb.15169
+* industrial development affects animals' ability to respond to both predictable and unpredictable change (https://www.nature.com/articles/s41559-022-01887-9)
+* effects of sudden deep snow in red desert on proghorn (https://www.sciencedirect.com/science/article/abs/pii/S0960982225002957)
+* https://www.nature.com/articles/s41467-023-37750-z
+* https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.4238
+* see data for snow and pronghorn paper (Aikens et al. 2025)
+* detecting regime shifts [in lakes: @bjorndahl_abrupt_2022]
+
+* smoothness of mu impacts smoothness and size of s2: depends on animals' ability to respond to, learn, and predict E(R)
+
+## Why not use machine learning? / Advantages of not using ML
+
+arguably, GAMs can be considered as a form of ML, but we are referring to NNs and other forms of black-box ML.
+
+\clearpage
+
 # Useful references
 
-* `{khroma}`: doi:10.5281/zenodo.1472077
+* `{khroma}`: @frerebeau_khroma_2024
 * @keith_predicting_2008
@@
 
+\clearpage
+
 # References

### 2025-04-14 analysis/011-mean-ndvi-hgam.R
fixed typos in comments

@@ -30,7 +30,7 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
   d0 <- readRDS('data/ndvi-global-15-day-average.rds')
   
   tictoc::tic()
-  #' `exactextractr::exact_extract()` fails due to bad  class for coords
+  #' `exactextractr::exact_extract()` fails due to bad class for coords
   #' takes 2.3 hours on EME linux
   d <- d0 %>%
     mutate(
@@ -134,7 +134,7 @@ if(file.exists('models/global-models/hbam-mean-ndvi-DATE.rds')) {
     knots = list(doy = c(0.5, 366.5)),
     drop.unused.levels = TRUE,
     discrete = TRUE,
-    samfrac = 0.001, # find intial guesses with a subset of the data
+    samfrac = 0.001, # find initial guesses with a subset of the data
     nthreads = future::availableCores(logical = FALSE) - 2,
     control = gam.control(trace = TRUE))
   

### 2025-04-14 analysis/012-var-ndvi-hgam.R
changed "ecoregions" to "biome", added `s(mu_hat)`

@@ -18,12 +18,13 @@ if(length(list.files('models/global-test', 'hbam-var-ndvi-*')) > 1) {
 } else {
   m <- bam(
     e_2 ~
-      wwf_ecoregion + # to avoid intercept shrinkage
+      biome + # to avoid intercept shrinkage
       s(poly_id, bs = 'mrf', xt = list(nb = nbs)) +
-      s(doy, wwf_ecoregion, bs = 'fs', xt = list(bs = 'cc'), k = 10) +
-      s(year, wwf_ecoregion, bs = 'fs', xt = list(bs = 'cr'), k = 10) +
-      ti(doy, year, wwf_ecoregion, bs = c('cc', 'cr', 're'), k = c(5, 5)) +
-      s(elevation_m, bs = 'cr', k = 5),
+      s(doy, biome, bs = 'fs', xt = list(bs = 'cc'), k = 10) +
+      s(year, biome, bs = 'fs', xt = list(bs = 'cr'), k = 10) +
+      ti(doy, year, biome, bs = c('cc', 'cr', 're'), k = c(5, 5)) +
+      s(elevation_m, bs = 'cr', k = 5) +
+      s(mu_hat, bs = 'cr', k = 5),
     family = gaussian(),
     data = d,
     method = 'fREML',

### 2025-04-14 writing/appendix-A.Rmd
re-arranged appendix figs based on refs in text

@@
 \noindent Figure A1 below illustrates the color palettes used in this appendix, for each type of vision (trichromatic, deuteranope, protanope, tritanope, and achomatic).
-Palette A was created by Masataka Okabe and Kei Ito [@okabe_color_2008], while the other palettes were created by Fabio Crameri [@crameri_geodynamic_2018; @crameri_scientific_2018].
+Palette A4 was created by Masataka Okabe and Kei Ito [@okabe_color_2008], while the other palettes were created by Fabio Crameri [@crameri_geodynamic_2018; @crameri_scientific_2018].
 All palettes below are available via the `khroma` package [@frerebeau_khroma_2024] for `R` [@r_core_team_r_2024].
 Figures A2-A5 show some of the data used for fitting the models for estimating mean NDVI and the variance around the mean.
-The code for generating the figures is available [on GitHub](https://github.com/QuantitativeEcologyLab/ndvi-stochasticity/blob/main/analysis/figures/input-data.R).
+The code for generating the figures is available on GitHub at [https://github.com/QuantitativeEcologyLab/ndvi-stochasticity/blob/main/analysis/figures/input-data.R](https://github.com/QuantitativeEcologyLab/ndvi-stochasticity/blob/main/analysis/figures/input-data.R).
 
@@
 
-```{r polygons, fig.cap="Map of the polygons for each ecoregion. Each polygon is colored randomly with one of 9 colors. Note that some neighboring polygons have the same color."}
-include_graphics('../figures/input-data/polygons.png')
+```{r n-rasters, fig.cap="Number of days with a raster within 15-day periods starting on January $1^{\\text{st}}$, for each year.
+Cells with less than 15 rasters indicate one or more missing rasters for that 15-day period, which the exception of cells near day 366, which have 6 days during leap years ($366 \\mod 15 = 6$) and 5 days otherwise ($365 \\mod 15 = 5$)."}
+include_graphics('../figures/input-data/n-rasters-time.png')
 ```
@@
 include_graphics('../figures/input-data/ecoregions.png')
+```
+
+```{r polygons, fig.cap="Map of the polygons for each ecoregion. Each polygon is colored randomly with one of 9 colors. Note that some neighboring polygons have the same color."}
+include_graphics('../figures/input-data/polygons.png')
 ```
@@
 
-```{r n-rasters, fig.cap="Number of days with a raster within 15-day periods starting on January $1^{\\text{st}}$, for each year.
-Cells with less than 15 rasters indicate one or more missing rasters for that 15-day period, which the exception of cells near day 366, which have 6 days during leap years ($366 \\mod 15 = 6$) and 5 days otherwise ($365 \\mod 15 = 5$)."}
-include_graphics('../figures/input-data/n-rasters-time.png')
-```
-
 \clearpage

### 2025-04-14 analysis/figures/010-input-data.R

@@ -6,6 +6,7 @@ library('cowplot')   # for fancy plots in grids
 library('tidyterra') # for plotting spatRasters with ggplot2
 library('khroma')    # for colorblind-friendly palettes
 source('analysis/figures/000-default-ggplot-theme.R')
+pal_polys <- c(color('okabeito')(8), 'grey')
 
 eco <- read_sf('data/ecoregions/ecoregions-polygons.shp') %>%
   filter(WWF_MHTNAM != 'Inland Water') %>%
@@ -72,7 +73,6 @@ if(FALSE) {
 
 # individual polygons ----
 n_distinct(eco$poly_id)
-pal_polys <- c(color('okabeito')(8), 'grey')
 plot_scheme_colorblind(pal_polys)
 ggsave('polygons-palette.png', path = 'figures/input-data',
        width = 6, height = 4, units = 'in', dpi = 300, bg = 'white')
@@ -196,15 +196,15 @@ plot_pal <- function(pal) {
 plot_pal(pal_polys)
 
 plot_grid(NULL,
-          plot_pal(pal_polys),
+          plot_pal(rev(color('lapaz')(1e3))),
           NULL,
           plot_pal(color('discreterainbow')(n_distinct(eco$WWF_MHTNAM))),
           NULL,
-          plot_pal(color('lajolla')(1e3)),
+          plot_pal(pal_polys),
           NULL,
-          # plot_pal(color('davos')(1e3)),
+          plot_pal(color('lajolla')(1e3)),
           # NULL,
-          plot_pal(rev(color('lapaz')(1e3))),
+          # plot_pal(color('davos')(1e3)),
           label_x = 0, ncol = 4,  rel_widths = c(0.075, 1),
           labels = c('A2.', '', 'A3.', '', 'A4.', '', 'A5.'))
 

### 2025-04-14 writing/manuscript.Rmd
started writing methods

@@
 
-\noindent Developments in remote sensing since the 1970s have greatly expanded our ability to quantify and monitor lanscapes using satellite imagery
+\noindent Developments in remote sensing since the 1970s have greatly expanded our ability to quantify and monitor landscapes using satellite imagery
 
@@
 * https://silvis.forest.wisc.edu/data/dhis/
+* we provide global rasters of the mean and variance in NDVI, averaged across 1981-2025
+* we provide a shiny app for people to calculate mean and variance without any knowledge of `R`
 
@@
 
-The models we fit
+We obtained NDVI data using the image composites from the AVHRR and VIIRS sensors [@vermote_noaa_2018; @vermote_noaa_2022].
+The code for downloading the data directly from the NOAA server is available at [https://github.com/QuantitativeEcologyLab/ndvi-stochasticity/blob/main/analysis/002-download-ndvi-rasters.R](https://github.com/QuantitativeEcologyLab/ndvi-stochasticity/blob/main/analysis/002-download-ndvi-rasters.R).
+The number of non-NA land raster cells was substantially larger than the maximum data frame size in `R` ($2.28 \times 10 ^{10} > 2^{31} - 1 \approx 0.21 \times 10^{10}$), so we reduced the dataset size by calculating 15-day averages of NDVI (see Fig.
+A2) and aggregating the averaged rasters by a factor of $2 \times 2$ with `terra::aggregate(fact = 2)`.
+Although this resulted in a reduction in sample size, the temporal averaging also reduced signal-related noise.
 
-### NDVI data
-
-* daily AVHRR-VIIRS
-* 15-day average and `aggregate(, 2)` to make dataset small enough to fit in a single data frame
+The predictor data for the models included: WWF biome [@olson_global_2002, see Fig.
+A3], ecoregion [the individual polygons from @olson_global_2002, see Fig.
+A4], integer day of year (1 to 366), integer year (1981 to 2025), and elevation above sea level (Fig.
+A5).
+We excluded all polygons that did not have any NDVI data (0.06% of land, 29% of the polygons, maximum area: $452~\text{km}^2 \approx (21~\text{km})^2$).
+We distinguished between biomes in the northern and southern hemisphere to allow for different seasonal trends between hemispheres.
+The 15-day averages allowed us to have non-repeating values of day of year across years, since 15 is not a factor of 365 or 366.
+We downloaded the global digital elevation model using the `get_elev_raster()` function from the `elevatr` package for `R` [v.
+0.99.0, @hollister_elevatr_2023] with a resolution of 0.076 degrees.
 
@@
 # approximate estimate of the number of unique DOY values
-length(unique(seq(1, 365.25 * (2025 - 1981), by = 15) %% c(365, 365, 365, 366)))
+length(unique(lubridate::yday(as.Date('1981-09-01') + seq(0, 15800, by = 15))))
 ```
-
-### Predictor data
-
-Maps are in Appendix A.
 
@@
 To reduce modeling fitting times with negligible losses to model accuracy, we used the `bam()` function with fast REstricted Marginal Likelihood (`method = fREML`) and covariate discretization (`discrete = TRUE`).
-The model structure is available below.
+See @wood_generalized_2015 and @wood_generalized_2017-1 for more information.
+Ideally, NDVI should be modeled using beta location-scale models (after the linear transformation $Y^*=\frac{Y + 1}{2}$) to account for the fact that: (1) NDVI is bounded between -1 and 1, and (2) the variance in NDVI is dependent on the mean (and vice-versa), since ecosystems with very low NDVI (e.g., rock, ice, or concrete) or very high NDVI (e.g., dense forest) tend to have lower variance in NDVI.
+However, preliminary tests showed that fitting times for beta models were prohibitive, especially for beta location-scale models.
+Additionally, Gaussian models fit substantially faster and provided very similar spatialtemporal estimates of mean NDVI (see the `analysis/000-sardinia-test.R` script in the GitHub repository).
+The model for the mean NDVI had the structure below:
 
-```{r, eval=FALSE}
-# the model for the mean
+\singlespacing
+
+\scriptsize
+
+```{r m_mean, eval=FALSE, echo=TRUE}
+m_mean <- bam(
+  ndvi_15_day_mean ~
+    biome + # to avoid intercept shrinkage
+    s(poly_id, bs = 'mrf', xt = list(nb = nbs)) +
+    s(doy, biome, bs = 'fs', xt = list(bs = 'cc'), k = 10) +
+    s(year, biome, bs = 'fs', xt = list(bs = 'cr'), k = 10) +
+    ti(doy, year, biome, bs = c('cc', 'cr', 're'), k = c(5, 5)) +
+    s(elevation_m, bs = 'cr', k = 5),
+  family = gaussian(),
+  data = d,
+  method = 'fREML',
+  knots = list(doy = c(0.5, 366.5)),
+  drop.unused.levels = TRUE,
+  discrete = TRUE,
+  samfrac = 0.001, # find initial guesses with a subset of the data
+  nthreads = future::availableCores(logical = FALSE) - 2,
+  control = gam.control(trace = TRUE))
 ```
+
+\normalsize
+
+\doublespacing
+
+Biome-specific fixed-effect intercepts allowed to account for the differences in mean NDVI throught biomes without coefficient shrinkage (i.e., reversion towards the global mean).
+A Markov Random Field of the ecoregions [`bs = mrf`, see page ***240?*** of @wood_generalized_2017] estimated smoothed, ecoregion-specific deviations from the biome-level spatial means.
+Two factor-smooth interaction terms (`bs = 'fs'`) accounted for biome-specific seasonal and yearly trends.
+Seasonal terms were made cyclical by using cyclical cubic splines (`xt = list(bs = 'cc')`), which are continuous up to and including second derivative at the edge knots (0.5 and 366.5, which correspond to 00:00 of January 1^st^ and 24:00 of December 31^st^, respectively).
+A tensor product interaction smooth for each biome estimated the change in seasonal trends over the years.
+Finally, a smooth of elevation above sea level accounted for the effect of altitude.
+See @pedersen_hierarchical_2019 for more information on hierarchical modeling and the use and interpretation of factor smooth interaction terms.
+
+***will we subtract the average residual from each pixel?***
+
+A second HGAM estimated the variance in NDVI around the mean estimated by the model above.
+The model had an identical structure to the model for the mean, with the exception that (1) the response variable was the squared residual from the first model, such that the model estimated the mean squared residual (i.e., the variance) for a given point in time and space, and (2) a smooth term of the estimated mean NDVI.
+The model is available below:
+
+\singlespacing
+
+\scriptsize
+
+```{r m_var, eval=FALSE, echo=TRUE}
+m_var <- bam(
+  e_2 ~
+    biome +
+    s(poly_id, bs = 'mrf', xt = list(nb = nbs)) +
+    s(doy, biome, bs = 'fs', xt = list(bs = 'cc'), k = 10) +
+    s(year, biome, bs = 'fs', xt = list(bs = 'cr'), k = 10) +
+    ti(doy, year, biome, bs = c('cc', 'cr', 're'), k = c(5, 5)) +
+    s(elevation_m, bs = 'cr', k = 5) +
+    s(mu_hat, bs = 'cr', k = 5),
+  family = gaussian(),
+  data = d,
+  method = 'fREML',
+  knots = list(doy = c(0.5, 366.5)),
+  drop.unused.levels = TRUE,
+  discrete = TRUE,
+  samfrac = 0.001,
+  nthreads = future::availableCores(logical = FALSE) - 2,
+  control = gam.control(trace = TRUE))
+```
+
+\normalsize
+
+\doublespacing
 
@@
 hex plots of mean and var vs common metrics:
+
 * temperature
@@
 * https://silvis.forest.wisc.edu/globalwui/
-* https://silvis.forest.wisc.edu/data/dhis/:
-  * NDVI16 (Normalized Difference Vegetation Index)
-  * EVI16 (Enhanced Vegetation Index)
-  * FPAR8 (Fraction absorbed Photosynthetically Active Radiation)
-  * LAI8 (Leaf Area Index)
-  * GPP8 (Gross Primary Productivity)
+* https://silvis.forest.wisc.edu/data/dhis/: NDVI16 (Normalized Difference Vegetation Index), EVI16 (Enhanced Vegetation Index), FPAR8 (Fraction absorbed Photosynthetically Active Radiation), LAI8 (Leaf Area Index), GPP8 (Gross Primary Productivity)
 * https://silvis.forest.wisc.edu/data/eurasia-trends/
@@
 
-* green wave surfing
+* green wave surfing: change in the mean, failure to surf: variance
 * migration routes are key habitat (Ortega et al., in prep), especially in stochastic environments
@@
 
-arguably, GAMs can be considered as a form of ML, but we are referring to NNs and other forms of black-box ML.
+arguably, GAMs can be considered as a form of machine learning, but we are referring to neural networks and other forms of black-box ML.
 

### 2025-04-14 writing/manuscript.Rmd
added 2 refs

@@
 
-We represent NDVI using a modified version of Fabio Crameri's divergent *bukavu* palette [@crameri_geodynamic_2018], which has high-contrast for deuteranope and protanope vision.
+We represent NDVI using a modified version of Fabio Crameri's divergent *bukavu* palette [@crameri_geodynamic_2018; @crameri_misuse_2020], which has high-contrast for deuteranope and protanope vision.
 For NDVI values between 0 and 1 (the right half of the palette), the colors also have sufficient contrast for the colors to be distinguishable by both tritanope and achomatic vision.
@@
 * @mezzini_how_2025
+* **TO READ:** Site fidelity as a maladaptive behavior in the Anthropocene: https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/fee.2456
 

### 2025-04-14 analysis/020-hex-plots.R
created some test hexplots using Rekha's estimates

@@ -1,0 +1,76 @@
+library('terra')   # for rasters
+library('dplyr')   # for data wrangling
+library('ggplot2') # for figures
+library('khroma')  # for colorblind-friendly color scale
+library('cowplot') # for ggplot plots in grids
+source('analysis/figures/000-default-ggplot-theme.R')
+
+r_mu <- rast('H:/GitHub/stochasticity_parks/Outputs/mean_predNDVI_raster.nc')
+r_s2 <- rast('H:/GitHub/stochasticity_parks/Outputs/var_residuals_raster.nc')
+r_dhi <- rast('data/other-rasters/dhi-data/dhi_ndvi_2015.tif')
+r_hfi <- rast('data/other-rasters/hfi-layers/ml_hfi_v1_2019.nc')
+
+layout(t(1:2))
+plot(r_mu)
+plot(r_s2)
+layout(1)
+plot(r_dhi)
+plot(r_hfi)
+
+d <- as.data.frame(r_s2, xy = TRUE) %>%
+  rename(s2 = 3) %>%
+  mutate(.,
+         mu = extract(r_mu, select(., x, y), xy = FALSE)[, 2],
+         hfi = extract(r_hfi, select(., x, y))[, 2]) %>%
+  bind_cols(.,
+            extract(r_dhi, select(., x, y)) %>%
+              select(! ID) %>%
+              rename(dhi_cumulative = dhi_ndvi_2015_1,
+                     dhi_min = dhi_ndvi_2015_2,
+                     dhi_seasonal = dhi_ndvi_2015_3)) %>%
+  as_tibble()
+d
+
+YLIMS <- c(NA, ceiling(max(d$s2) * 100) / 100)
+ZLIMS <- log(c(1, 150))
+
+plot_hex <- function(variable, bins = 50, xlab, .data = d) {
+  names(.data)[colnames(.data) == variable] <- 'variable'
+  
+  #' not adding `geom_smooth()` because it's too slow with large data sets
+  .data %>%
+    filter(, ! is.na(variable)) %>%
+    ggplot() +
+    geom_hex(aes(variable, s2, fill = log10(after_stat(count))),
+             color = 'black', bins = bins, linewidth = 0.25)+
+    scale_fill_acton(name = 'Count (log scale)', range = c(0, 1),
+                     limits = ZLIMS, reverse = TRUE,
+                     labels = \(.lab) round(exp(.lab))) +
[…]

### 2025-04-15 analysis/figures/020-hex-plots.R
moved to `figures` folder

@@ -46,7 +46,7 @@ plot_hex <- function(variable, bins = 50, xlab, .data = d) {
     scale_fill_acton(name = 'Count (log scale)', range = c(0, 1),
                      limits = ZLIMS, reverse = TRUE,
                      labels = \(.lab) round(exp(.lab))) +
-    labs(x = xlab, y = 'Variance in NDVI') +
+    labs(x = xlab, y = 'DENVar') +
     ylim(YLIMS) +
     theme(legend.position = 'none') +
     #' for a safety check for fill limits, `ZLIMS`

### 2025-04-15 analysis/011-mean-ndvi-hgam.R
rm polys w/o data; changed ecoregions to biomes

@@ -15,7 +15,7 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
 } else {
   # resolution is (0.05 degrees)^2 = (3 degree minutes)^2
   # create raster of ecoregion
-  r_eco <- rast('data/ecoregions/wwf-ecoregions.tif')
+  r_biomes <- rast('data/ecoregions/wwf-ecoregions.tif')
   
   # create raster of polygon ID
   r_poly_id <- rast('data/ecoregions/ecoregion-polygon-id.tif')
@@ -37,11 +37,11 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
       year = lubridate::year(central_date),
       doy = lubridate::yday(central_date),
       poly_id = extract(r_poly_id, data.frame(x, y))[, 2], # 1 is pixel ID
-      wwf_ecoregion = extract(r_eco, data.frame(x, y))[, 2],
+      biome = extract(r_biomes, data.frame(x, y))[, 2],
       # distance_coast_m = extract(r_dist, data.frame(x, y)))[, 2],
       elevation_m = extract(r_elev, data.frame(x, y))[, 2]) %>%
     # convert strings to factors while separating the two hemispheres
-    mutate(wwf_ecoregion = paste(wwf_ecoregion, if_else(y < 0, 'S', 'N')) %>%
+    mutate(biome = paste(biome, if_else(y < 0, 'S', 'N')) %>%
              factor())
   
   d <- na.omit(d)
@@ -57,7 +57,9 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
       points(y ~ x, ., col = 'red', pch = 19)
   }
   
-  d <- mutate(d, elevation_m = if_else(elevation_m < 0, 0, elevation_m))
+  d <- mutate(d, elevation_m = if_else(elevation_m < 0, 0, elevation_m)) %>%
+    mutate(doy = as.integer(doy),
+           year = as.integer(year))
   
   saveRDS(d, 'data/hbam-ndvi-data.rds')
   tictoc::toc()
@@ -70,7 +72,11 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
 # create list of neighbors ----
 # drop polygons with no data or neighbors
 ecoregions$in_data <- ecoregions$poly_id %in% unique(d$poly_id)
-filter(ecoregions, ! in_data & n_neigh == 0)
+
+filter(ecoregions, ! in_data)
[…]

### 2025-04-15 analysis/010-test-hgams.R
fixed typos in comments

@@ -3,8 +3,8 @@
 library('dplyr')   # for data wrangling
 library('purrr')   # for functional programming
 library('sf')      # for shapefiles
-library('mgcv')    # for Genralized Additive Models
-library('sf')      # for simple fleature objects
+library('mgcv')    # for Generalized Additive Models
+library('sf')      # for simple feature objects
 library('terra')   # for rasters
 library('elevatr') # for extracting elevation
 library('gratia')  # for plotting GAMs

### 2025-04-15 analysis/010-test-hgams.R
fit test for sos smooth

@@ -112,7 +112,7 @@ if(FALSE) {
     geom_point()
 }
 
-ecoregions <- filter(ecoregions, in_data | n_neigh > 0)
+ecoregions <- filter(ecoregions, in_data)
 
 nbs <- spdep::poly2nb(pl = ecoregions,
                       row.names = ecoregions$poly_id, # so that names match IDs
@@ -135,7 +135,7 @@ add_nb(p1 = 'poly 8038', p2 = 'poly 8040', add = TRUE)
 add_nb(p1 = 'poly 7987', p2 = 'poly 7988', add = TRUE)
 add_nb(p1 = 'poly 7987', p2 = 'poly 8040', add = TRUE)
 add_nb(p1 = 'poly 7988', p2 = 'poly 8038', add = TRUE)
-add_nb('poly 8980', 'poly 8987', add = TRUE)
+add_nb(p1 = 'poly 8980', p2 = 'poly 8987', add = TRUE)
 layout(1)
 
 all(names(nbs) == ecoregions$poly_id)
@@ -289,17 +289,72 @@ d <- mutate(d,
             mu_hat = mu_hat + mean_e)
 
 # fit the model for the variance
-m_var <- bam(ndvi_15_day_mean ~
-            s(elevation_m, bs = 'cr', k = 5) +
-            s(doy, wwf_ecoregion, bs = 'fs', xt = list(bs = 'cc'), k = 10) +
-            s(poly_id, bs = 'mrf', xt = list(nb = nbs)),
-          family = gaussian(),
-          data = d,
-          method = 'fREML',
-          knots = list(doy = c(0.5, 366.5)),
-          drop.unused.levels = FALSE,
-          discrete = TRUE,
-          nthreads = 10,
-          control = gam.control(trace = TRUE))
+m_var <- bam(e_2 ~
+               s(elevation_m, bs = 'cr', k = 5) +
+               s(doy, wwf_ecoregion, bs = 'fs', xt = list(bs = 'cc'), k = 10) +
+               s(poly_id, bs = 'mrf', xt = list(nb = nbs)),
+             family = gaussian(),
+             data = d,
+             method = 'fREML',
+             knots = list(doy = c(0.5, 366.5)),
+             drop.unused.levels = FALSE,
+             discrete = TRUE,
+             nthreads = 10,
+             control = gam.control(trace = TRUE))
 draw(m_var, rug = FALSE, select = 1:2)
[…]

### 2025-04-23 analysis/figures/presentation-figures/wiggliness-mean-var.R
started presentation for lab update

@@ -1,0 +1,67 @@
+library('dplyr') # for data wrangling
+library('tidyr') # for data wrangling
+library('mgcv') # for modeling
+library('ggplot2') # for plotting
+source('analysis/figures/000-default-ggplot-theme.R')
+
+d <- tibble(x = runif(1e3),
+            mu = sinpi(sqrt(x) * 10 -0.5) + 3 * sqrt(x),
+            s2 = cospi(sqrt(x) - 0.5),
+            y = rnorm(n = length(x), mean = mu, sd = sqrt(s2)))
+
+layout(1:3); plot(mu ~ x, d); plot(s2 ~ x, d); plot(y ~ x, d); layout(1)
+
+m_1 <- gam(list(y ~ s(x), ~ s(x)), family = gaulss(), data = d, method = 'REML')
+l_inv_s <- function(link_pred) (exp(link_pred) + 0.01)
+
+tibble(x = seq(0, 1, by = 1e-3)) %>%
+         bind_cols(predict(m_1, newdata = ., type = 'link', se.fit = TRUE) %>%
+                     as.data.frame() %>%
+                     rename(mu_est = 1, s2_est = 2,
+                            mu_se = 3, s2_se = 4)) %>%
+  pivot_longer(-x, names_sep = '_', names_to = c('par', 'measure')) %>%
+  pivot_wider(names_from = measure, values_from = value) %>%
+  mutate(lab = if_else(par == 'mu', 'Mean', 'Variance'),
+         lwr = if_else(par == 'mu', est - 1.96 * se,
+                       l_inv_s(est - 1.96 * se)),
+         upr = if_else(par == 'mu', est + 1.96 * se,
+                       l_inv_s(est + 1.96 * se)),
+         est = if_else(par == 'mu', est, l_inv_s(est))) %>%
+  ggplot() +
+  facet_wrap(~ lab, scales = 'free', strip.position = 'left', ncol = 1) +
+  geom_point(aes(x, y), mutate(d, lab = 'Mean'), alpha = 0.1) +
+  geom_ribbon(aes(x, ymin = lwr, ymax = upr), alpha = 0.2) +
+  geom_line(aes(x, est), lwd = 1) +
+  ylab(NULL) +
+  theme(strip.placement = 'outside', strip.background = element_blank(),
+        strip.text = element_text(size = 12))
+  
+ggsave('figures/presentation-figures/mean-var-wiggly.png',
[…]

### 2025-04-23 writing/manuscript.Rmd
added refs and text to discussion

@@
 We estimate spatiotemporal trends in mean NDVI and the variance around the mean using Hierarchical Generalized Additive Models, which provide a flexible yet transparent model structure, unlike machine learning approaches such as neural netwoks.
-We show that DENVar can be used as a reliable proxy of environmental stochasticity, strongly correlates with ***XXX***, and can be used to test many hepotheses related to environmental stochasitcity, such as forage unpredictability, regime shifts, ...
+We show that DENVar can be used as a reliable proxy of environmental stochasticity, strongly correlates with ***XXX***, and can be used to test many hepotheses related to environmental stochasitcity, such as forage stochasticity, regime shifts, ...
 .
@@
 
-\noindent Developments in remote sensing since the 1970s have greatly expanded our ability to quantify and monitor landscapes using satellite imagery
+\noindent Since its early uses in the 1970s, the Normalized Difference Vegetation Index (NDVI) has greatly expanded our ability to quantify and monitor landscapes using satellite imagery
 
@@
 However, preliminary tests showed that fitting times for beta models were prohibitive, especially for beta location-scale models.
-Additionally, Gaussian models fit substantially faster and provided very similar spatialtemporal estimates of mean NDVI (see the `analysis/000-sardinia-test.R` script in the GitHub repository).
-The model for the mean NDVI had the structure below:
+In contrast, Gaussian models fit substantially faster and provided very similar spatialtemporal estimates of mean NDVI (see the `analysis/000-sardinia-test.R` script in the GitHub repository).
+The global model for the mean NDVI had the structure below:
 
@@
 
-\scriptsize
+\small
 
@@
 ```{r m_var, eval=FALSE, echo=TRUE}
+#' *CHANGE TO SOS BASIS*
 m_var <- bam(
@@
 
-output of `draw()`
+## General results
 
-hex plots of mean and var vs common metrics:
+* two Orthographic projections of DENVar: Northern and southern hemispheres on spring solstice
+
+## Correlations with other metrics
+
+```{r hex-plots, fig.cap="Hex plots of the relationship between the global Dynamic Estimate of NDVI Variance (DENVar) and common ecological metricts.
+The fill of the hexagonal cells represents the log-10-transformed number of points within the cell.
+Rowwise, the variables are: mean NDVI (as estimated in this manuscript); the machine-learning Human Footprint Index (ml-HFI) of Keys, Barnes \\& Carter (2021); the 2015 cumulative, minimum, and seasonal range in the Dynamic Habitat Index of XXX, (F)."}
+knitr::include_graphics('../figures/hexplots-canada-test.png')
+```
 
@@
 * gross primary productivity
-* hfi
 * max animal weight
@@
 * https://silvis.forest.wisc.edu/data/dhis/: NDVI16 (Normalized Difference Vegetation Index), EVI16 (Enhanced Vegetation Index), FPAR8 (Fraction absorbed Photosynthetically Active Radiation), LAI8 (Leaf Area Index), GPP8 (Gross Primary Productivity)
-* https://silvis.forest.wisc.edu/data/eurasia-trends/
+
+## How to obtain the data
+
+### Static rasters of mean and variance (1981-2025)
+
+* for GIS users (tif)
+
+### Shiny app for obtaining values
+
+* get values for specific coordinates: 1981-2025 average
+* get values for specific coordinates and dates (upload as a CSV, download as a CSV)
+* screenshot of the shiny app
 
@@
 
-* green wave surfing: change in the mean, failure to surf: variance
+We have presented a new spatiotemporally dynamic metric of environmental stochasticity, which we have called the Dynamic Estimate of NDVI Variance, DENVar.
+In this section, we discuss the interpretation of DENVar in greater depth, and we argue the advantages and limitations of the statistical methods used.
+We conclude by suggsting some use cases of DENVar within animal movement ecology, landscape ecology, and phenology, but we invite readers to consider how DENVar may be of use in their field and adjacent disciplines.
+
+## Why use complicated statistics instead of machine learning?
+
+\noindent Admittedly, the title of this section is misleading for two reasons.
+Firstly, some may argue that Generalized Additive Models are a form of machine-learning (here abbreviated as ML), but we intend to contrast GAMs with more "black-box" types of machine learning, such as neural networks and random forests.
+The second reason for which the title is misleading is that it implies that machine learning is not complicated, or at least that it is less complicated than the models we used.
+However, we argue that while traditional statistical models may seem more intimidating to some, their complexity is what allows analysts to produce interpretable results and measures of uncertainty [@wood_generalized_2017].
+And while the theoretical basis for GAMs can discouraging for those who do not a strong background in mathematics or statistics, outputs from GAMs can be presented in an accessible manner even to those without a strong quantitative background through the use of well-designed figures, especially in a Bayesian context (when the aim is to estimate a parameter).
+In contrast, while ML methods will often give understandable results, it may not be possible to discern *why* such results were "chosen" by the model.
+Instead, GAMs produce flexible yet interpretable estimated effects that can be represented with relatively simple figures, even if the mathematical formula(e) for the model may be complicated.
+
+Ideally, a model should estimate or explore the relationships among variables such that results can be interpreted in a transparent and unambiguous manner, such that readers can scrutinize the underlying assumptions, methods, and decisions.
+We recognize that the assessing GAMs requires a certain level of expertise, but we hope the readers can appreciate the option to assess the GAMs more easily than a series of consecutive, weighted regressions (as in the case of neural networks).
+In contrast, GAMs learn from data with no initial starting assumptions other than user-imposed limitations of wiggliness, which are clearly outlined in each basis size of the models' terms.
+The structure and wuggliness of GAMS can thus be grossly designed based on the system's complexity and structure while allowing the data to inform the model trends, wiggliness, and coefficient size.
+
+## Interpreting DENVar: the importance of scale
+
+* smoothness of mean impacts smoothness and size of variance: depends on scale of interest (e.g., animals' ability to respond to, learn, and predict changes in the mean).
+Changes not attributed to trends in the mean are stochastic
+* DENVar is a good metric as long as the spatiotemporal scale of interest is similar or larger to that of DENVar.
+May not work well for small-scale processes such as the movement dynamics of a small rodent, the phenology of a small orchard when spatiotemporal heterogeneity is finer than DENVar's resolution
+
+## Environmental stochasticity and animal movement ecology
+
+The field of animal movement ecology has long shown that animals alter their movement in response to resource abundance or availability [@burt_territoriality_1943; @southwood_habitat_1977; @charnov_optimal_1976; @broekman_environmental_2024].
+More recently, movement ecologists have started focusing more on resource stochasticity [e.g., @stephens_optimal_1982; @rizzuto_forage_2021], but it is worth distinguishing between resource stochasticity and seasonality [@nilsen_can_2005].
+Seasonality is generally best interpreted as a cyclical change in the mean that many animals can learn to predict, as in the case of green-wave surfing [@middleton_green-wave_2018; @geremia_migrating_2019].
+In contrast, environmental stochasticity refers to unpredictable changes, such as forest fires [@ref], sudden snow storms [@aikens_pronghorn_2025], and unusual phenology [@ref].
+Animals also respond to stochastic changes [@mezzini_how_2025], but their ability to adapt depends on behavioral plasticity [@steixner-kumar_strategies_2020; @rickbeil_plasticity_2019], and memory [@polansky_elucidating_2015; @abrahms_memory_2019; @riotte-lambert_environmental_2020; @falcon-cortes_hierarchical_2021; @ranc_memory_2022].
+
+@botero_evolutionary_2015 tipping points
+
+@polazzo_measuring_2024
+
+extreme events, black swans [@anderson_black-swan_2017; @logares_black_2012]
+
 * migration routes are key habitat (Ortega et al., in prep), especially in stochastic environments
-* @sawyer_migratory_2019
+* can fail to adapt to changes in conditions @sawyer_migratory_2019
 * https://www.sciencedirect.com/science/article/pii/S0960982220308484
@@
 * https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.4238
-* see data for snow and pronghorn paper (Aikens et al. 2025)
 * detecting regime shifts [in lakes: @bjorndahl_abrupt_2022]
 
-* smoothness of mu impacts smoothness and size of s2: depends on animals' ability to respond to, learn, and predict E(R)
+## Environmental stochasticity and population dynamics
 
-## Why not use machine learning? / Advantages of not using ML
+read @lande_risks_1993
 
-arguably, GAMs can be considered as a form of machine learning, but we are referring to neural networks and other forms of black-box ML.
+\noindent Environmental stochasticity reduces habitats' energetic balance and carrying capacity [@chevin_adaptation_2010] as individuals struggle to rely on unpredictable resources and increase the instability of populations and communities [@lande_risks_1993].
+Additionally, stochastic environments reduce the ability of specialist species to adapt and evolve while selecting for more generalist species [@levins_evolution_1974].
+Consequently, stochastic environments tend to have lower species diversity [@marcus].
+
+## Environmental stochasticity and fire ecology
+
+\noindent Recent work by @devries has suggested that NDVI stochasticity is an important metric for evaluating forest recovery after large fires, while @collinson has suggested that variance in NDVI is an important predictor of the occurrence of forest fires due to ignition by lightning.
+
+## Environmental stochasticity and phenology
+
+* @jonzen_rapid_2006
+* effects can be compounded by cumulative effects (e.g., loss of species diversity, see @wolf_flowering_2017)
+* climate change affects plant phenology, but changes in situ may not match experimental estimates [@wolkovich_warming_2012]
+* reproduction of painted lady butterfly [@stefanescu_timing_2021]
+* @wessling_seasonal_2018
 
@@
 
-* `{khroma}`: @frerebeau_khroma_2024
 * @keith_predicting_2008
@@
 * @nilsen_can_2005
-* @stephens_optimal_1982
 * @merkle_large_2016
@@
 * @mezzini_how_2025
-* **TO READ:** Site fidelity as a maladaptive behavior in the Anthropocene: https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/fee.2456
+
+TO READ:
+* Site fidelity as a maladaptive behavior in the Anthropocene: https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/fee.2456
+* Pettorelli N., Vik J.O., Mysterud A., Gaillard J.-M., Tucker C.J.
+& Stenseth N.Chr.
+(2005).
+Using the satellite-derived NDVI to assess ecological responses to environmental change.
+Trends in Ecology & Evolution 20, 503–510.
+https://doi.org/10.1016/j.tree.2005.05.011
 

### 2025-04-23 functions/get_legend.R
added custom function to extract ggplot legend

@@ -1,0 +1,8 @@
+library('ggplot2')
+library('cowplot')
+
+#' `get_legend()` for `{cowplot}` version 1.1.3 returns an empty plot 
+get_legend <- function(.plot) {
+  get_plot_component(.plot + theme(legend.position = 'top'),
+                     pattern = 'guide-box-top', return_all = TRUE)
+}

### 2025-04-24 analysis/010-test-hgams.R
fit sos test GAMs

@@ -311,12 +311,13 @@ d <- d %>%
 
 # k = 1e3: fits in ~ 30 seconds
 # k = 2e3: fits in ~ 653 seconds
-m_sos <- bam(ndvi_15_day_mean ~ s(x, y, bs = 'sos', k = 2e3),
-             family = gaussian(),
-             data = d,
-             method = 'fREML',
-             discrete = TRUE,
-             control = gam.control(trace = TRUE))
+m_sos <- bam(ndvi_15_day_mean ~ s(y, x, bs = 'sos', k = 2e3), # s(lat,long)
+  family = gaussian(),
+  data = d,
+  method = 'fREML',
+  discrete = TRUE,
+  control = gam.control(trace = TRUE))
+saveRDS(m_sos, 'models/global-test/test-mean-gam-sos-only.rds')
 
 png('figures/test-hgams/test-hgam-sos.png', width = 7.5, height = 10,
     units = 'in', bg = 'white', res = 300)
@@ -358,3 +359,15 @@ m_re <- bam(ndvi_15_day_mean ~ s(poly_id, bs = 're'),
 tictoc::toc()
 (1 - m_re$deviance / m_re$null.deviance) * 100
 m_re$df.null - m_re$df.residual
+
+# test for model of the variance
+d <- mutate(d, e2 = resid(m_sos)^2)
+
+m_sos_var <- bam(
+  e2 ~ s(y, x, bs = 'sos', k = 2e3), # s(lat, long)
+  family = gaussian(),
+  data = d,
+  method = 'fREML',
+  discrete = TRUE,
+  control = gam.control(trace = TRUE))
+saveRDS(m_sos_var, 'models/global-test/test-var-gam-sos-only.rds')

### 2025-04-24 analysis/figures/020-hex-plots.R
switched to a single plot with `facet_wrap()`

@@ -4,24 +4,29 @@ library('ggplot2') # for figures
 library('khroma')  # for colorblind-friendly color scale
 library('cowplot') # for ggplot plots in grids
 source('analysis/figures/000-default-ggplot-theme.R')
+source('functions/get_legend.R')
 
-r_mu <- rast('H:/GitHub/stochasticity_parks/Outputs/mean_predNDVI_raster.nc')
-r_s2 <- rast('H:/GitHub/stochasticity_parks/Outputs/var_residuals_raster.nc')
+m_mu <- readRDS('models/global-test/test-mean-gam-sos-only.rds')
+m_s2 <- readRDS('models/global-test/test-var-gam-sos-only.rds')
+
 r_dhi <- rast('data/other-rasters/dhi-data/dhi_ndvi_2015.tif')
 r_hfi <- rast('data/other-rasters/hfi-layers/ml_hfi_v1_2019.nc')
+r_rich <- rast('data/other-rasters/iucn-red-list-spp-richness/Combined_SR_2024.tif') %>%
+  project('EPSG:4326') # all others are lat/long
+crs(r_dhi, proj = TRUE) == crs(r_hfi, proj = TRUE) &
+  crs(r_hfi, proj = TRUE) == crs(r_rich, proj = TRUE)
 
-layout(t(1:2))
-plot(r_mu)
-plot(r_s2)
-layout(1)
+# check rasters
 plot(r_dhi)
 plot(r_hfi)
+plot(r_rich)
 
-d <- as.data.frame(r_s2, xy = TRUE) %>%
-  rename(s2 = 3) %>%
-  mutate(.,
-         mu = extract(r_mu, select(., x, y), xy = FALSE)[, 2],
-         hfi = extract(r_hfi, select(., x, y))[, 2]) %>%
+#' *will need to change the data later*
+d <- m_mu$model %>%
+  mutate(mu_hat = fitted(m_mu),
+         s2_hat = fitted(m_s2),
+         hfi = extract(r_hfi, select(., x, y))[, 2],
+         richness = extract(r_rich, select(., x, y))[, 2]) %>%
   bind_cols(.,
             extract(r_dhi, select(., x, y)) %>%
               select(! ID) %>%
@@ -31,46 +36,88 @@ d <- as.data.frame(r_s2, xy = TRUE) %>%
   as_tibble()
 d
 
-YLIMS <- c(NA, ceiling(max(d$s2) * 100) / 100)
[…]

### 2025-04-24 analysis/011-mean-ndvi-hgam.R
switched spatial smooths to sos basis

@@ -10,8 +10,9 @@ source('functions/add_nb.R')
 ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
 
 # set up the data for the model
+#' *need to re-run to fix elevations*
 if(file.exists('data/hbam-ndvi-data.rds')) {
-  d <- readRDS('data/hbam-ndvi-data.rds')
+  d <- readRDS('data/hbam-ndvi-data-only-years-multiples-of-5.rds') #' *change*
 } else {
   # resolution is (0.05 degrees)^2 = (3 degree minutes)^2
   # create raster of ecoregion
@@ -27,7 +28,8 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
   r_dist <- rast('data/distance-from-coast-m.tif')
   
   # takes a few minutes to import
-  d0 <- readRDS('data/ndvi-global-15-day-average.rds')
+  d0 <- readRDS('data/ndvi-global-15-day-average.rds') %>%
+    filter(lubridate::year(central_date) %% 5 == 0) #' *DELETE*
   
   tictoc::tic()
   #' `exactextractr::exact_extract()` fails due to bad class for coords
@@ -53,7 +55,7 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
   if(FALSE) {
     plot(st_geometry(ecoregions))
     d %>%
-      filter(elevation_m < -20) %>%
+      filter(elevation_m < -0) %>%
       points(y ~ x, ., col = 'red', pch = 19)
   }
   
@@ -69,73 +71,75 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
   gc()
 }
 
-# create list of neighbors ----
-# drop polygons with no data or neighbors
-ecoregions$in_data <- ecoregions$poly_id %in% unique(d$poly_id)
-
-filter(ecoregions, ! in_data)
-mean(! ecoregions$in_data)
-with(ecoregions, sum((! in_data) * area_km2) / sum(area_km2))
-max(ecoregions$area_km2[ecoregions$in_data])
-
-if(FALSE) {
-  ggplot(ecoregions, aes(log10(area_km2), in_data)) + geom_point()
-  
-  ggplot(ecoregions, aes(log10(area_km2), n_neigh)) +
-    facet_wrap(~ in_data) +
-    geom_point()
-}
-
[…]

### 2025-04-24 analysis/012-var-ndvi-hgam.R

@@ -4,68 +4,164 @@ library('mgcv')    # for Genralized Additive Models
 library('sf')      # for simple fleature objects
 library('terra')   # for rasters
 library('gratia')  # for plotting GAMs
+library('ggplot2') # for fancy plots
 source('analysis/figures/000-default-ggplot-theme.R')
 
-ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
+# ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
 
 # import the data for the model
-d <- readRDS('data/hbam-var-ndvi-data.rds')
-nbs <- readRDS('data/ecoregions/poly-nbs-global.rds')
+d <- readRDS('data/hbam-var-ndvi-data-mod-5-no-res-2025-04-21-THINNED-50.rds')
+
+hist(d$mu_hat)
+quantile(d$mu_hat, 0.001)
+# d <- d %>%
+#   mutate(mu_hat = if_else(mu_hat < quantile(mu_hat, 0.001),
+#                           quantile(mu_hat, 0.001), mu_hat),
+#          mu_hat = if_else(mu_hat > 1, 1, mu_hat))
+# nbs <- readRDS('data/ecoregions/poly-nbs-global.rds')
 
 if(length(list.files('models/global-test', 'hbam-var-ndvi-*')) > 1) {
   DATE <- '2025-04-XX'
-  m <- readRDS(paste0('models/global-test/hbam-var-ndvi-', DATE, '.rds'))
+  m_s2 <- readRDS(paste0('models/global-test/hbam-var-ndvi-', DATE, '.rds'))
 } else {
-  m <- bam(
+  m_s2 <- bam(
     e_2 ~
-      biome + # to avoid intercept shrinkage
-      s(poly_id, bs = 'mrf', xt = list(nb = nbs)) +
-      s(doy, biome, bs = 'fs', xt = list(bs = 'cc'), k = 10) +
-      s(year, biome, bs = 'fs', xt = list(bs = 'cr'), k = 10) +
-      ti(doy, year, biome, bs = c('cc', 'cr', 're'), k = c(5, 5)) +
+      s(doy, bs = 'cc', k = 10) +
+      s(year, bs = 'cr', k = 9) +
+      s(y, x, bs = 'sos', k = 1e3) + # sos uses lat, long
+      ti(doy, y, x, bs = c('cc', 'sos'), d = c(1, 2), k = c(10, 250)) +
[…]

### 2025-04-24 analysis/011-mean-ndvi-hgam.R
updated models in `readRDS()`; fixed filename typo

@@ -119,14 +119,14 @@ if(file.exists('data/hbam-ndvi-data.rds')) {
 #'   layout(1)
 #'   saveRDS(nbs, 'data/ecoregions/poly-nbs-global.rds')
 #' }
-
-# ensure names match the factor levels and the unique values
-# d <- mutate(d, poly_id = factor(poly_id, levels = names(nbs)))
-all.equal(sort(names(nbs)), sort(levels(d$poly_id)))
-all(levels(d$poly_id) %in% unique(d$poly_id))
+#' 
+#' # ensure names match the factor levels and the unique values
+#' d <- mutate(d, poly_id = factor(poly_id, levels = names(nbs)))
+#' all.equal(sort(names(nbs)), sort(levels(d$poly_id)))
+#' all(levels(d$poly_id) %in% unique(d$poly_id))
 
 if(file.exists('models/global-models/hbam-mean-ndvi-DATE.rds')) {
-  m_mu <- readRDS('models/global-models/hbam-mean-ndvi-NO-MRF-NO-TI-THINNED-DATA-2025-04-14.rds')
+  m_mu <- readRDS('models/global-models/hbam-mean-ndvi-sos-mod-5-no-res-2025-04-21-THINNED-50.rds')
 } else {
   DATE <- paste0('mod-5-no-res-', format(Sys.Date(), '%Y-%m-%d'))
   

### 2025-04-24 analysis/012-var-ndvi-hgam.R

@@ -7,8 +7,6 @@ library('gratia')  # for plotting GAMs
 library('ggplot2') # for fancy plots
 source('analysis/figures/000-default-ggplot-theme.R')
 
-# ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
-
 # import the data for the model
 d <- readRDS('data/hbam-var-ndvi-data-mod-5-no-res-2025-04-21-THINNED-50.rds')
 
@@ -22,7 +20,7 @@ quantile(d$mu_hat, 0.001)
 
 if(length(list.files('models/global-test', 'hbam-var-ndvi-*')) > 1) {
   DATE <- '2025-04-XX'
-  m_s2 <- readRDS(paste0('models/global-test/hbam-var-ndvi-', DATE, '.rds'))
+  m_s2 <- readRDS(paste0('models/global-models/hbam-var-ndvi-sos-mod-5-no-res-2025-04-21-THINNED-50.rds'))
 } else {
   m_s2 <- bam(
     e_2 ~
@@ -44,7 +42,7 @@ if(length(list.files('models/global-test', 'hbam-var-ndvi-*')) > 1) {
     nthreads = future::availableCores(logical = FALSE) - 2,
     control = gam.control(trace = TRUE))
   
-  saveRDS(m_s2, paste0('models/global-models/hbam-mean-ndvi-sos-', DATE, '.rds'))
+  saveRDS(m_s2, paste0('models/global-models/hbam-var-ndvi-sos-', DATE, '.rds'))
 }
 
 100 * (1 - m_s2$deviance / m_s2$null.deviance)

### 2025-04-24 analysis/figures/020-hex-plots.R
created global hex plot

@@ -1,3 +1,4 @@
+library('sf')      # for simple features
 library('terra')   # for rasters
 library('dplyr')   # for data wrangling
 library('ggplot2') # for figures
@@ -6,25 +7,32 @@ library('cowplot') # for ggplot plots in grids
 source('analysis/figures/000-default-ggplot-theme.R')
 source('functions/get_legend.R')
 
-m_mu <- readRDS('models/global-test/test-mean-gam-sos-only.rds')
-m_s2 <- readRDS('models/global-test/test-var-gam-sos-only.rds')
-
+r_s2 <- rast('data/output/var-ndvi-raster-mod-5-no-res-2025-04-21-THINNED-50.tif')
+ecoregions <- read_sf('data/ecoregions/ecoregions-polygons.shp') %>%
+  # drop polygons not in data
+  filter(poly_id %in% names(readRDS('data/ecoregions/poly-nbs-global.rds'))) %>%
+  vect() %>% # convert to spatVect
+  project(crs(r_s2))
+r_s2 <- crop(r_s2, ecoregions, mask = TRUE)
+r_mu <- rast('data/output/mean-ndvi-raster-mod-5-no-res-2025-04-21-THINNED-50.tif')
 r_dhi <- rast('data/other-rasters/dhi-data/dhi_ndvi_2015.tif')
-r_hfi <- rast('data/other-rasters/hfi-layers/ml_hfi_v1_2019.nc')
+r_hfi <- rast('data/other-rasters/hfi-layers/ml_hfi_v1_2019.nc') # no data on greenland or arctic
 r_rich <- rast('data/other-rasters/iucn-red-list-spp-richness/Combined_SR_2024.tif') %>%
-  project('EPSG:4326') # all others are lat/long
-crs(r_dhi, proj = TRUE) == crs(r_hfi, proj = TRUE) &
-  crs(r_hfi, proj = TRUE) == crs(r_rich, proj = TRUE)
+  project(crs(r_s2))
+n_distinct(sapply(list(r_s2, r_mu, r_dhi, r_hfi, r_rich), \(x) crs(x, proj = TRUE)))
 
 # check rasters
-plot(r_dhi)
-plot(r_hfi)
-plot(r_rich)
+if(FALSE) {
+  plot(r_s2)
+  plot(r_mu)
+  plot(r_dhi)
+  plot(r_hfi)
+  plot(r_rich)
+}
 
[…]

### 2025-04-24 analysis/figures/presentation-figures/single-raster.R
made fig of example AVHRR/VIIRS raster

@@ -1,0 +1,27 @@
+library('terra')  # for rasters
+library('sf')     # for simple features
+library('ggplot2') # for fancy plots
+library('dplyr')  # for data wrangling
+source('analysis/figures/000-default-ggplot-theme.R')
+r <- rast("H:/GitHub/ndvi-stochasticity/data/avhrr-viirs-ndvi/raster-files/AVHRR-Land_v005_AVH13C1_NOAA-07_19810701_c20170609165932.nc", lyr = 'NDVI')
+eco <- st_read('data/ecoregions/ecoregions-polygons.shp') %>%
+  filter(WWF_MHTNAM != 'Inland Water') %>%
+  st_geometry()
+
+plot(r)
+plot(eco, add = TRUE)
+
+d <- as.data.frame(r, xy = TRUE)
+nrow(d) / 1e6
+
+p <- ggplot() +
+  geom_sf(data = eco, color = 'black', fill = 'grey30', lwd = 0.1) +
+  geom_raster(aes(x, y, fill = NDVI), d) +
+  geom_sf(data = eco, color = 'black', fill = 'transparent', lwd = 0.1) +
+  scale_x_continuous(NULL, expand = c(0, 0)) +
+  ylab(NULL) +
+  scale_fill_gradientn(colours = ndvi_pal, limits = c(-1, 1)) +
+  theme(legend.position = 'top')
+
+ggsave('figures/ndvi-1981-07-01-raw-raster.png',
+       p, width = 8, height = 4, units = 'in', bg = 'white', dpi = 300)

### 2025-04-24 analysis/figures/020-hex-plots.R
made hexplot fig wider; fixed missing comma

@@ -66,12 +66,11 @@ d %>%
   scale_fill_iridescent(
     name = expression(paste(bold('Count (log'), bold(''['10']),
                             bold(' scale)'))), range = c(0, 1),
-    reverse = FALSE
-    labels = \(.x) 10^.x) +
+    reverse = FALSE, labels = \(.x) 10^.x) +
   labs(y = NULL, x = 'DENVar') +
   theme(legend.position = 'top', strip.background = element_blank(),
         strip.placement = 'outside', legend.key.width = unit(0.5, 'in'),
         strip.text = element_text(size = rel(1)))
 
 ggsave('figures/hexplots-global-test.png',
-       width = 12, height = 10, units = 'in', dpi = 600, bg = 'white')
+       width = 16, height = 10, units = 'in', dpi = 600, bg = 'white')

### 2025-04-24 analysis/figures/presentation-figures/single-raster.R
removed great lakes from background polygon

@@ -6,6 +6,7 @@ source('analysis/figures/000-default-ggplot-theme.R')
 r <- rast("H:/GitHub/ndvi-stochasticity/data/avhrr-viirs-ndvi/raster-files/AVHRR-Land_v005_AVH13C1_NOAA-07_19810701_c20170609165932.nc", lyr = 'NDVI')
 eco <- st_read('data/ecoregions/ecoregions-polygons.shp') %>%
   filter(WWF_MHTNAM != 'Inland Water') %>%
+  filter(ECO_NAME != 'Great Lakes') %>%
   st_geometry()
 
 plot(r)

### 2025-04-24 analysis/figures/020-hex-plots.R
changed ml-HFI to HFP of Mu et al., decr fig size

@@ -16,7 +16,7 @@ ecoregions <- read_sf('data/ecoregions/ecoregions-polygons.shp') %>%
 r_s2 <- crop(r_s2, ecoregions, mask = TRUE)
 r_mu <- rast('data/output/mean-ndvi-raster-mod-5-no-res-2025-04-21-THINNED-50.tif')
 r_dhi <- rast('data/other-rasters/dhi-data/dhi_ndvi_2015.tif')
-r_hfi <- rast('data/other-rasters/hfi-layers/ml_hfi_v1_2019.nc') # no data on greenland or arctic
+r_hfi <- rast('data/other-rasters/hfp_2021_100m_v1-2_cog.tif')
 r_rich <- rast('data/other-rasters/iucn-red-list-spp-richness/Combined_SR_2024.tif') %>%
   project(crs(r_s2))
 n_distinct(sapply(list(r_s2, r_mu, r_dhi, r_hfi, r_rich), \(x) crs(x, proj = TRUE)))
@@ -30,13 +30,24 @@ if(FALSE) {
   plot(r_rich)
 }
 
+get_values <- function(rst, pts) {
+  extract(rst, select(pts, x, y) %>%
+            vect(geom = c('x', 'y')) %>%
+            set.crs('EPSG:4326') %>%
+            project(crs(rst))) %>%
+    pull(2)
+}
+
 d <- as.data.frame(r_mu, xy = TRUE) %>%
   mutate(.,
-         s2_hat = extract(r_s2, select(., x, y))[, 2],
-         hfi = extract(r_hfi, select(., x, y))[, 2],
-         richness = extract(r_rich, select(., x, y))[, 2]) %>%
+         s2_hat = get_values(r_s2, .),
+         hfi = get_values(r_hfi, .) / 1e3, # scale back to [0, 50]
+         richness = get_values(r_rich, .)) %>%
   bind_cols(.,
-            extract(r_dhi, select(., x, y)) %>%
+            extract(r_dhi, select(., x, y) %>%
+                      vect(geom = c('x', 'y')) %>%
+                      set.crs('EPSG:4326') %>%
+                      project(crs(r_dhi))) %>%
               select(! ID) %>%
               rename(dhi_cumulative = dhi_ndvi_2015_1,
                      dhi_min = dhi_ndvi_2015_2,
[…]

### 2025-04-25 analysis/figures/020-hex-plots.R
updated description for HFP; need to remake fig

@@ -16,7 +16,7 @@ ecoregions <- read_sf('data/ecoregions/ecoregions-polygons.shp') %>%
 r_s2 <- crop(r_s2, ecoregions, mask = TRUE)
 r_mu <- rast('data/output/mean-ndvi-raster-mod-5-no-res-2025-04-21-THINNED-50.tif')
 r_dhi <- rast('data/other-rasters/dhi-data/dhi_ndvi_2015.tif')
-r_hfi <- rast('data/other-rasters/hfp_2021_100m_v1-2_cog.tif')
+r_hfi <- rast('H:/GitHub/ndvi-stochasticity/data/other-rasters/hfp_2021_100m_v1-2_cog.tif')
 r_rich <- rast('data/other-rasters/iucn-red-list-spp-richness/Combined_SR_2024.tif') %>%
   project(crs(r_s2))
 n_distinct(sapply(list(r_s2, r_mu, r_dhi, r_hfi, r_rich), \(x) crs(x, proj = TRUE)))
@@ -66,7 +66,7 @@ d %>%
   tidyr::pivot_longer(c(mu_hat, hfi:dhi_seasonal), names_to = 'variable',
                       values_to = 'value') %>%
   mutate(lab = case_when(variable == 'mu_hat' ~ 'Mean NDVI',
-                         variable == 'hfi' ~ '2021 Human Footprint',
+                         variable == 'hfi' ~ '2017-2021 Human Footprint',
                          variable == 'richness' ~ 'Species richness',
                          variable == 'dhi_cumulative' ~ 'Cumulative DHI',
                          variable == 'dhi_min' ~ 'Minimum DHI',

### 2025-04-25 analysis/figures/020-hex-plots.R
fixed ylab of hex plot, added hist of DENVar

@@ -61,12 +61,19 @@ if(FALSE) {
     plot(y ~ x, .)
 }
 
+ggplot(d, aes(s2_hat)) +
+  geom_histogram(fill = 'grey', color = 'black', binwidth = 0.005,
+                 center = 0.0025) +
+  labs(x = 'DENVar', y = 'Count')
+ggsave('figures/denvar-hist-global-test.png',
+       width = 5, height = 3, units = 'in', dpi = 600, bg = 'white')
+
 d %>%
   filter(mu_hat > -0.25) %>% #' *remove: there were 4 points mean < -0.25*
   tidyr::pivot_longer(c(mu_hat, hfi:dhi_seasonal), names_to = 'variable',
                       values_to = 'value') %>%
   mutate(lab = case_when(variable == 'mu_hat' ~ 'Mean NDVI',
-                         variable == 'hfi' ~ '2017-2021 Human Footprint',
+                         variable == 'hfi' ~ 'Human Footprint',
                          variable == 'richness' ~ 'Species richness',
                          variable == 'dhi_cumulative' ~ 'Cumulative DHI',
                          variable == 'dhi_min' ~ 'Minimum DHI',

### 2025-04-25 analysis/012-var-ndvi-hgam.R
added dashed line at y = 0 to hex plot

@@ -84,6 +84,7 @@ d %>%
   ggplot() +
   geom_hex(aes(mu_hat, s2_hat, fill = log10(after_stat(count))),
            color = 'black', bins = 100, linewidth = 0.1, na.rm = TRUE) +
+  geom_hline(yintercept = 0, color = 'grey') +
   labs(x = 'Estimated mean NDVI', y = 'DENVar') +
   scale_fill_iridescent(
     name = expression(paste(bold('Count (log'), bold(''['10']),

### 2025-04-25 analysis/figures/021-denvar-biomes.R
made figures of histograms for each biome

@@ -1,0 +1,177 @@
+library('sf')      # for shapefiles
+library('terra')   # for rasters
+library('dplyr')   # for data wrangling
+library('ggplot2') # for fancy plots
+library('cowplot') # for fancy plots in grids
+source('analysis/figures/000-default-ggplot-theme.R')
+
+biomes <- rast('data/ecoregions/wwf-ecoregions.tif')
+plot(biomes)
+biomes_sf <- st_read('data/ecoregions/ecoregions-polygons.shp')
+unique(biomes_sf$WWF_MHTNAM)
+mu <- rast('data/output/mean-ndvi-raster-mod-5-no-res-2025-04-21-THINNED-50.tif')
+denvar <- rast('data/output/var-ndvi-raster-mod-5-no-res-2025-04-21-THINNED-50.tif')
+plot(denvar)
+
+d <- as.data.frame(mu, xy = TRUE) %>%
+  mutate(.,
+         s2_hat = as.data.frame(denvar)[[1]],
+         i = extract(biomes, select(., x, y))[, 2],
+         biome = sort(unique(ecoregions$WWF_MHTNAM))[i + 1]) %>%
+  filter(biome != 'Inland Water') %>%
+  mutate(biome = factor(biome, levels = c(
+    'Rock and Ice',
+    'Tundra',
+    'Boreal Forests/Taiga',
+    'Montane Grasslands and Shrublands',
+    'Temperate Grasslands, Savannas and Shrublands',
+    'Temperate Broadleaf and Mixed Forests',
+    'Temperate Conifer Forests',
+    'Mediterranean Forests, Woodlands and Scrub',
+    'Mangroves',
+    'Tropical and Subtropical Moist Broadleaf Forests',
+    'Tropical and Subtropical Dry Broadleaf Forests',
+    'Tropical and Subtropical Coniferous Forests',
+    'Tropical and Subtropical Grasslands, Savannas and Shrublands',
+    'Flooded Grasslands and Savannas',
+    'Deserts and Xeric Shrublands'))) %>%
+  as_tibble()
+
+biomes_sf <- biomes_sf %>%
+  filter(WWF_MHTNAM != 'Inland Water') %>%
+  mutate(WWF_MHTNAM = factor(WWF_MHTNAM, levels = levels(d$biome)))
+
+biome_pal <- color('discreterainbow')(n_distinct(d$biome))
+
+ggplot(d, aes(x, y, fill = biome)) +
[…]

### 2025-04-25 analysis/figures/010-input-data.R
changed "ecoregions" to "biomes"

@@ -91,16 +91,16 @@ ggsave('figures/input-data/polygons.png', p_poly,
 # ecoregions ----
 n_distinct(eco$WWF_MHTNAM)
 
-p_eco <-
+p_biome <-
   ggplot(eco) +
   geom_sf(aes(fill = WWF_MHTNAM), color = 'black', lwd = .05) +
   geom_hline(yintercept = 0, color = 'black', lwd = 0.1, lty = 'dashed') +
-  scale_fill_discreterainbow(name = 'Ecoregion') +
+  scale_fill_discreterainbow(name = 'Biome') +
   scale_x_continuous(expand = c(0, 0)) +
   theme(legend.position = 'top', legend.text = element_text(size = 4.5))
 
-ggsave('figures/input-data/ecoregions.png', p_eco,
-       width = 10.1, height = 6, units = 'in', dpi = 600, bg = 'white')
+ggsave('figures/input-data/ecoregions.png', p_biome,
+       width = 10.1, height = 6, units = 'in', dpi = 1e3, bg = 'white')
 
 # elevation ----
 r_elev <- rast('data/elev-raster.tif') %>%

### 2025-05-14 analysis/001-download-ndvi-rasters.R
re-downloaded data and cleaned download script

@@ -1,3 +1,4 @@
+# last run on 2025-05-13
 library('rvest') # for harvesting web pages; see https://rvest.tidyverse.org/
 library('ncdf4') # for nc rasters
 library('dplyr') # for data wrangling
@@ -8,7 +9,7 @@ plan(multisession, workers = ncores)
 
 # main url and years to find all folders in the parent directory
 url_main <- 'https://www.ncei.noaa.gov/data/land-normalized-difference-vegetation-index/access/'
-years <- 1981:2025 #' years to download; *downloading up to 2025-05-06*
+years <- 1981:2025 #' years to download; *downloading up to 2025-05-07*
 
 # find file names for each raster
 fn_tib <- tibble(
@@ -28,7 +29,7 @@ tail(fn_tib)
 # download the files for each year
 sum(grepl('preliminary', fn_tib$fn)) # check if any rasters are preliminary
 options('timeout') # in seconds
-options(timeout = 60 * 10) # increase timeout to 10 minutes
+options(timeout = 60 * 20) # increase timeout to 20 minutes
 options('timeout') # in seconds
 
 DIR <- '~/bigdrive/shared/avhrr-viirs-ndvi-raster-files/'
@@ -49,12 +50,10 @@ output <-
   }, .progress = TRUE)
 table(grepl('Downloaded', output))
 
-#' **HERE** --------------------------------------------------------------------
-
 #' There's a few days that have no raster associated with them, but it's
 #' generally rare. We have 8 2-day gaps, 3 3-day gaps, 2 4-day gaps, and
 #' one 16-day gap. The gaps range from 1982 to 2024, and the longest one
-#' is in summer of 2022.
+#' is in summer of 2022 (see `gaps` below).
 
 # check files
 fn_dates <- fn_tib %>%
@@ -65,19 +64,15 @@ fn_dates <- fn_tib %>%
   as.Date(format = '%Y%m%d')
 
 main_dates <-
-  list.files('data/avhrr-viirs-ndvi/raster-files', pattern = '*.nc',
-             recursive = FALSE) %>%
+  list.files(DIR, pattern = '*.nc', recursive = FALSE) %>%
   substr(.,
          nchar(.) - nchar('YYYYMMDD_cYYYYmmddHHMMSS.nc') + 1,
          nchar(.) - nchar('_cYYYYmmddHHMMSS.nc')) %>%
   as.Date(format = '%Y%m%d')
 
-# ensure date ranges are the same
 range(fn_dates)
-all(range(fn_dates) == range(main_dates))
-
[…]

### 2025-05-14 analysis/002-download-ndvi-rasters.R
renamed scripts

@@ -1,106 +1,0 @@
-library('rvest') # for harvesting web pages; see https://rvest.tidyverse.org/
-library('ncdf4') # for nc rasters
-library('dplyr') # for data wrangling
-library('purrr') # for functional programming
-library('furrr') # for parallel functional programming
-ncores <- min(50, future::availableCores(logical = FALSE) / 2)
-plan(multisession, workers = ncores)
-
-# main url and years to find all folders in the parent directory
-url_main <- 'https://www.ncei.noaa.gov/data/land-normalized-difference-vegetation-index/access/'
-years <- 1981:2025 #' years to download; *downloading up to 2025-03-16*
-
-# find file names for each raster
-fn_tib <- tibble(
-  years = years,
-  urls = future_map(years, function(y) {
-    url_y <- paste0(url_main, y, '/')
-    daysfiles <- XML::htmlParse(httr::GET(url_y)) %>%
-      XML::xpathSApply(., path = '//a', XML::xmlGetAttr, 'href')
-    
-    # identify and subset all of the links with '.nc' in the name
-    return(tibble(fn = daysfiles[grepl('\\.nc', daysfiles)]))
-  })) %>%
-  tidyr::unnest(urls)
-fn_tib
-tail(fn_tib, 10)
-sum(grepl('preliminary', fn_tib$fn)) # check if any rasters are preliminary
-
-# download the files for each year
-options('timeout') # in seconds
-options(timeout = 60 * 10) # increase timeout to 10 minutes
-options('timeout') # in seconds
-
-DIR <- paste0('//home/mezzinis/H/GitHub/ndvi-stochasticity/',
-              'data/avhrr-viirs-ndvi/raster-files/')
-if(! dir.exists(DIR)) stop('DIR does not exist!')
-plan() # check plan
-
-output <-
-  future_map2_chr(fn_tib$years, fn_tib$fn, function(.year, .filename) {
-    if(! file.exists(paste0(DIR, .filename))) {
-      download.file(
-        url = paste0(url_main, .year, '/', .filename),
-        #' cannot save directly to `//home/shared` because it needs sudo access
-        destfile = paste0(DIR, .filename), method = 'libcurl', quiet = TRUE)
-      return(paste('Downloaded', .filename))
-    } else {
-      warning(paste('Did not download', .filename))
-      return(paste('Did not download', .filename))
[…]

### 2025-05-14 analysis/002-sardinia-test.R

@@ -359,6 +359,7 @@ if(file.exists('models/sardinia-test/beta-gam-mrf.rds')) {
 #' requires `LAPACK = TRUE` in `base::qr.default()`, which can only be done with
 #' `assignInNamespace()` because `fixInNamespace()` can't edit the function
 #' fitting time is >> 144 hours = 6 days on EME Linux (3 newtons in 6 days)
+#' uses too much RAM (> 68.4 GB) and swap memory (total: > 716 GB)
 if(FALSE) {
   system.time(
     m_betals <- gam(list(

### 2025-05-16 writing/manuscript.Rmd
added some refs and example fig for mean and var

@@
     institute: [braes, biol]
+  - name: Gavin L. Simpson
+    institute: [au]
   - name: Michael J. Noonan
@@
   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - au: Department of Animal & Veterinary Sciences, Aarhus University, Denmark.
   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
@@
 subparagraph: true # needed for \usepackage[small]{titlesec}
+urlcolor: cyan
 output:
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, out.width = '100%', fig.align = 'center')
+knitr::opts_chunk$set(echo = FALSE, out.width = '100%', fig.align = 'center', message = FALSE, warning = FALSE, comment = '#')
 ```
@@
 
-\noindent Since its early uses in the 1970s, the Normalized Difference Vegetation Index (NDVI) has greatly expanded our ability to quantify and monitor landscapes using satellite imagery
-
-- NDVI is an estimate of greenness, NPP, forage availability
-- NDVI has many uses and is often convenient because it's free, well-established, and
+\noindent Since its early uses in the 1970s, the Normalized Difference Vegetation Index (NDVI) has greatly expanded our ability to quantify and monitor landscapes using remotely sensed imagery [@pettorelli_using_2005; @pettorelli_normalized_2011; @anyamba_historical_2012].
+NDVI is an estimate of a landscape's greenness, and it often a good proxy of primary productivity, forage availability, and vegetative cover [@pettorelli_normalized_2011; @merkle_large_2016].
+However, as NDVI a proxy, it is not a direct measurement of such variables, and it should be interpreted carefully.
+NDVI is prone to loss of sensitivity in highly saturated (i.e., very green) environments [@matsushita_sensitivity_2007], and estimates can vary substantially across sensors [@huang_commentary_2021].
+Still, NDVI is often used because it is free, well-established, and available at a wide variety of spatiotemporal scales [@ref].
 
 <!-- successes of NDVI: pettorelli, Aikens -->
+
+NDVI has been particularly useful when studying the movement behavior of herbivorous mammals, such as when testing the relationship between resource availability and range size [@ref]
+
+NDVI has also been used to test the "green wave hypothesis", which was first developed by @ref in relation to migratory birds and states that animals will time their migration with the early green-up phases of forage growth to maximize forage nutrient content.
 
@@
 
+scale in remote sensing and GIS: @goodchild_scale_2023
+
 <!-- difficulties in estimating variance -->
 
-* variance is hard to calculate, as it depends on a correct estimate of the mean
+* variance is hard to calculate, as it depends on a the estimated mean
 * paper on estimating the asymptote of environmental variance globally, but variance for a closed system must converge towards a finite value [@bachmaier_variogram_2011; @fleming_fine-scale_2014]
+
+```{r fig-smooth-wiggly, cache=TRUE, fig.cap="Examples of smooth and wiggly estimated trends in simulated NDVI data (points).
+Smooth estimates of the mean (top left) result in wiggly and large variance estimates (bottom left), as short-term trends in the data are assumed to be extreme events.
+In contrast, wiggly (i.e., flexible) estimates of the mean (top right) result in smooth and small variance estimates (bottom right), since short-term trends in the data are assumed to be trends in the mean.
+The data is the same in both columns.", fig.width=9, fig.height=4.5}
+library('dplyr')   # for data wrangling
+library('tidyr')   # for data wrangling
+library('mgcv')    # for modeling
+library('ggplot2') # for plotting
+source('../analysis/figures/000-default-ggplot-theme.R')
+
+# OU process given samples times, reversion time, standard deviation 
+ou <- function(t, theta, sigma){
+  n <- length(t) # number of samples
+  dw  <- rnorm(n = n, mean = 0, sd = sqrt(sigma)) # noise between steps
+
+  x <- numeric(n)
+  x[1] <- 0 # set starting point
+  for (i in 2:n) {
+    x[i] <- (1 - theta) * x[i-1] + dw[i - 1]
+  }
+  return(x)
+}
+
+set.seed(3)
+d <- tibble(t = seq(0, 1, length.out = 400),
+            mu = (sinpi(sqrt(t + 0.5) * 10 - 0.5) + 1 * sqrt(t)) / 50 + 0.3,
+            s2 = cospi(sqrt(t) - 0.5) * 0.001^2,
+            noise = ou(t = t, theta = 0.2, sigma = sqrt(s2)),
+            y = mu + noise)
+
+if(FALSE) {
+  ggplot(d) +
+    geom_line(aes(t, mu), color = 'red') +
+    geom_line(aes(t, y), alpha = 0.5)
+}
+
+m_wiggly <- gam(list(y ~ s(t, k = 30, bs = 'ad'), ~ s(t)),
+                family = gaulss(), data = d, method = 'REML')
+m_smooth <- gam(list(y ~ s(t, k = 10), ~ s(t)),
+                family = gaulss(), data = d, method = 'REML')
+
+get_preds <- function(mod, prefix = 'mod_') {
+  predict(mod, newdata = tibble(t = seq(0, 1, by = 1e-3)),
+          type = 'response', se.fit = FALSE) %>%
+    as.data.frame() %>%
+    transmute(Mean_est = V1, Variance_est = 1 / V2^2) %>%
+    rename_with(.fn = \(.x) paste0(prefix, .x), .cols = 1:2)
+}
+
+bind_cols(tibble(t = seq(0, 1, by = 1e-3)),
+          get_preds(m_wiggly, 'Wiggly_'), get_preds(m_smooth, 'Smooth_')) %>%
+  pivot_longer(-t, names_sep = '_',
+               names_to = c('model', 'lab', 'measure')) %>%
+  pivot_wider(names_from = measure, values_from = value) %>%
+  mutate(model = paste(model, 'estimate')) %>%
+  ggplot() +
+  geom_point(aes(t, y), tibble(t = 0, y = 0, lab = 'Variance'),
+             color = 'transparent') +
+  facet_grid(lab ~ model, scales = 'free', switch = 'y', ) +
+  geom_point(aes(t, y), mutate(d, lab = 'Mean'), alpha = 0.1) +
+  geom_line(aes(t, est), linewidth = 0.75) +
+  ylab(NULL) +
+  scale_x_continuous('Time', breaks = NULL) +
+  theme(strip.placement = 'outside', strip.background.y = element_blank(),
+        strip.text = element_text(size = 10),
+        axis.title.x = element_text(size = 10))
+```
 
@@
 
-```{r, eval=FALSE}
+```{r n_doys, eval=FALSE}
 # approximate estimate of the number of unique DOY values
@@
 
+figures:
+
+- plot grid of mean over space, `s(year)`, `s(doy)`, `s(elev)`
+- plot grid of DENVar over space, `s(year)`, `s(doy)`, `s(elev)`, `s(mean)`
+- histograms of DENVar around flat map of biomes (change colors)
+- hex plots of DENVAr with other variables (add temperature, annual precip, extreme weather events, fires, invasive species, max body size, disease occurrence, ...)
+
 ## General results
@@
 Rowwise, the variables are: mean NDVI (as estimated in this manuscript); the machine-learning Human Footprint Index (ml-HFI) of Keys, Barnes \\& Carter (2021); the 2015 cumulative, minimum, and seasonal range in the Dynamic Habitat Index of XXX, (F)."}
-knitr::include_graphics('../figures/hexplots-canada-test.png')
+knitr::include_graphics('../figures/hexplots-global-test.png')
 ```
@@
 
+supplementary material:
+
+- hex plots of pred vs obs for mean and var
+- histograms of mean around flat map of biomes (change colors)
+
+\clearpage
+
 # References

### 2025-05-16 analysis/002-sardinia-test.R
removed unnecessary code and objects

@@ -24,31 +24,27 @@ assignInNamespace(x = 'qr.default',   # replace `base::qr.default()`
                   value = qr.default, # with the local version sourced above
                   ns = 'base')        # specify the base package
 
-# the bounding box for sardinia (large enough to include all coast)
-sardinia_bbox <- tibble(x = c(7.6, 10.4), y = c(38.8, 41.3)) %>%
-  st_as_sf(coords = c('x', 'y')) %>%
-  st_bbox() %>%
-  st_as_sfc() %>%
-  st_as_sf() %>%
-  st_set_crs('EPSG:4326')
-
 # shapefile of the world's terrestrial ecoregions
 sardinia <- read_sf('data/world-ecosystems/data/commondata/data0/tnc_terr_ecoregions.shp') %>%
   st_make_valid() %>%
-  st_intersection(sardinia_bbox) %>%
+  st_intersection(
+    # the bounding box for sardinia (large enough to include all coast)
+    tibble(x = c(7.6, 10.4), y = c(38.8, 41.3)) %>%
+      st_as_sf(coords = c('x', 'y')) %>%
+      st_bbox() %>%
+      st_as_sfc() %>%
+      st_as_sf() %>%
+      st_set_crs('EPSG:4326')) %>%
   st_cast('POLYGON') %>%
   st_as_sf() %>%
-  mutate(poly_id = paste('poly', 1:n()),
-         area_km2 = as.numeric(st_area(.)) / 1e6)
-hist(sardinia$area_km2)
-
-sardinia_geom <- st_geometry(sardinia)
+  st_geometry()
 
 # import ndvi data (downloaded from the NASA website) ----
 if(file.exists('data/sardinia-test/sardinia-ndvi.rds')) {
   d <- readRDS('data/sardinia-test/sardinia-ndvi.rds')
 } else {
-  if(.Platform$OS.type != 'unix') stop('AVHRR/VIIRS rasters are on the H: Drive, and you may want to use nultiple cores.')
+  if(.Platform$OS.type != 'unix')
+    stop('AVHRR/VIIRS rasters are on the H: Drive, and you may want to use multiple cores.')
   future::availableCores(logical = FALSE)
   plan(multisession, workers = min(30, availableCores(logical = FALSE) - 2))
   d <-
@@ -58,8 +54,8 @@ if(file.exists('data/sardinia-test/sardinia-ndvi.rds')) {
[…]

### 2025-05-16 analysis/003-ecoregion-polygons.R
cleaned up script for ecoregion polys

@@ -1,0 +1,59 @@
+library('dplyr')   # for data wrangling
+library('tidyr')   # for data wrangling
+library('purrr')   # for data wrangling
+library('sf')      # for spatial data
+library('terra')   # for rasters
+library('spdep')   # for finding neighbors
+library('ggplot2') # for fancy plots
+sf_use_s2(FALSE)
+
+# get shapefile of each continent
+ecoregions <- st_read('data/world-ecosystems/data/commondata/data0/tnc_terr_ecoregions.shp') %>%
+  mutate(group = case_when(WWF_REALM %in% c('NA', 'NT') ~ 'americas',
+                           WWF_REALM %in% c('AT', 'PA', 'IM') ~ 'africa+eurasia',
+                           WWF_REALM %in% c('AA', 'OC') ~ 'islands',
+                           WWF_REALM == 'AN' ~ 'Antarctica',
+                           .default = 'unassigned')) %>%
+  select(group, ECO_NAME, ECO_NUM, WWF_REALM, WWF_MHTNAM) %>%
+  st_make_valid() %>% # to drop duplicate vertices
+  mutate(area_km2 = as.numeric(st_area(.)) / 1e6) %>%
+  filter(WWF_MHTNAM != 'Inland Water') %>% # rm 17 inland water polygons
+  st_transform(crs = crs(rast('data/avhrr-viirs-ndvi/raster-files/AVHRR-Land_v005_AVH13C1_NOAA-07_19810624_c20170610041337.nc',
+                              lyr = 'NDVI')))
+
+if(FALSE) { # some exploratory plots and checks
+  # each ecoregion has a unique name, but many ecoregions are multipolygons
+  unique(ecoregions$ECO_NAME)[1:10]
+  plot(st_geometry(ecoregions), col = factor(ecoregions$ECO_NAME),
+       axes = TRUE)
+  
+  # by group ----
+  # find total area by group
+  ecoregions %>%
+    st_drop_geometry() %>%
+    group_by(group) %>%
+    summarize(total = sum(area_km2)) %>%
+    mutate(total = total / sum(total))
+  
+  ggplot(ecoregions) +
+    facet_wrap(~ group) +
+    geom_sf(fill = 'red3', color = 'black') +
+    theme_classic()
+  
+  # by WWF realm ----
+  # find total area by WWF_REALM
+  ecoregions %>%
+    st_drop_geometry() %>%
+    group_by(WWF_REALM) %>%
[…]

### 2025-05-16 analysis/003-merging-rasters.R
cleaned up code for rast merge; need to choose res

@@ -1,211 +1,0 @@
-library('dplyr') # for data wrangling
-library('tidyr') # for data wrangling
-library('sf')    # for shapefiles
-library('terra') # for rasters
-library('purrr') # for functional programming
-
-file_names <-
-  list.files(path = 'data/avhrr-viirs-ndvi/raster-files/',
-             pattern = '.nc', full.names = TRUE, recursive = FALSE)
-length(file_names) / 365 # number of years
-sum(grepl('prelim', file_names)) # ensure no rasters are preliminary
-
-# check a few rasters
-if(FALSE) {
-  plot(rast(file_names[[1]]))
-  plot(rast(file_names[[2]])) # compare NDVI to QA
-  plot(rast(file_names[[3]]))
-  plot(rast(file_names[[4]]))
-  plot(rast(file_names[[100]]))
-  plot(rast(file_names[[200]]))
-} # looks like filtering based on QA has already been done
-
-# no clear connection between QA and NDVI; maybe data are already clean
-if(FALSE) {
-  r <- rast(file_names[2])
-  plot(values(r$QA), values(r$NDVI))
-  rm(r)
-}
-
-#' one AVHRR raster has ~6M cells, which means only about 330 global rasters fit
-#' in a data frame, since the max number of rows is (2^31 - 1) = 2e9
-ecoregions <- st_read('data/world-ecosystems/data/commondata/data0/tnc_terr_ecoregions.shp') %>%
-  filter(WWF_REALM2 != 'Antarctic') # drop Antarctica
-
-# all rasters have the same CRS
-# check crs for first, last, and a random sample of rasters
-c(file_names[1], sample(file_names, size = 100), file_names[length(file_names)]) %>%
-  map_chr(function(.fn) { # fast enough that it is not worth parallelizing
-    crs(rast(.fn), proj = TRUE)
-  }) %>%
-  unique()
-
-ecoregions <- st_transform(ecoregions, crs(rast(file_names[1]))) %>%
-  st_geometry() %>%
-  vect()
-
-# create a data frame of all dates
-dates <-
-  tibble(date = seq(as.Date('1981-06-24'), as.Date('2025-03-15'), by = 1),
-         julian = julian(date),
-         group = julian - (julian %% 15),
-         file_name = map_chr(date, function(.date) {
[…]

### 2025-05-16 analysis/004-merging-rasters.R

@@ -1,0 +1,290 @@
+library('dplyr') # for data wrangling
+library('tidyr') # for data wrangling
+library('sf')    # for shapefiles
+library('terra') # for rasters
+library('purrr') # for functional programming
+library('furrr') # for parallelized functional programming
+
+file_names <-
+  list.files(path = 'data/avhrr-viirs-ndvi/raster-files/',
+             pattern = '.nc', full.names = TRUE, recursive = FALSE)
+length(file_names) / 365 # approximate number of years of data
+
+# check a few rasters
+if(FALSE) {
+  plot(rast(file_names[[1]]))
+  plot(rast(file_names[[2]])) # compare NDVI to QA
+  plot(rast(file_names[[3]]))
+  plot(rast(file_names[[4]]))
+  plot(rast(file_names[[100]]))
+  plot(rast(file_names[[200]]))
+  
+  # looks like filtering based on QA has already been done
+  rast(file_names[[1]]) %>%
+    as.data.frame() %>%
+    slice_sample(n = 1e4) %>%
+    plot(is.na(NDVI) ~ QA, .)
+  
+  # no clear connection between QA and NDVI; maybe data are already clean
+  r <- rast(file_names[2])
+  plot(values(r$QA), values(r$NDVI))
+  rm(r)
+}
+
+# all rasters have the same CRS:
+# check crs for first, last, and a random sample of rasters
+file_names[c(1, length(file_names),
+             sample(length(file_names), size = 100))] %>%
+  map_chr(function(.fn) { # fast enough that it is not worth parallelizing
+    crs(rast(.fn))
+  }) %>%
+  unique()
+
+#' `ecoregions` uses same projection as first raster, `file_names[1]`
+ecoregions <- st_read('data/ecoregions/ecoregions-polygons.shp')
+
+# find number of cells per complete raster (i.e., assuming no NAs)
+max_size <- 2^31 - 1 # max data frame size in R
+n_cells <-
+  rast(file_names[1], lyr = 'QA') %>% # cells in a raster w no NAs
+  mask(ecoregions) %>%
+  not.na() %>%
+  values() %>%
+  sum()
+n_cells / 1e6 # million land cells per raster (including NAs) 
+
+# lat: 1 deg = ~110 km
+# long: 1 deg = 111.320*cos(lat) km
+tibble(
+  lat = seq(-90, 90, by = 30),
+  long = 0,
+  lat_deg_to_km = 110,
[…]

### 2025-05-16 analysis/004-ecoregion-neighbors.R
no longer using nbs based on ecoregion polygons

@@ -1,344 +1,0 @@
-library('dplyr')   # for data wrangling
-library('tidyr')   # for data wrangling
-library('purrr')   # for data wrangling
-library('sf')      # for spatial data
-library('terra')   # for rasters
-library('spdep')   # for finding neighbors
-library('ggplot2') # for fancy plots
-source('functions/add_nb.R')
-
-# get shapefile of each continent
-ecoregions <- st_read('data/world-ecosystems/data/commondata/data0/tnc_terr_ecoregions.shp') %>%
-  filter(WWF_REALM2 != 'Antarctic') %>% # drop Antarctica
-  #' **\/ for testing **
-  # filter(WWF_REALM == 'PA') %>%
-  # st_make_valid() %>%
-  # st_intersection(filter(world, name_long == 'Italy') %>%
-  #                   st_buffer(10e3) %>%
-  #                   st_bbox() %>%
-  #                   st_as_sfc() %>%
-  #                   st_as_sf() %>%
-  #                   st_set_crs('EPSG:4326')) %>%
-  #' **/\ for testing **
-  mutate(group = case_when(WWF_REALM %in% c('NA', 'NT') ~ 'americas',
-                           WWF_REALM %in% c('AT', 'PA', 'IM') ~ 'africa+eurasia',
-                           WWF_REALM %in% c('AA', 'OC') ~ 'islands',
-                           .default = 'unassigned')) %>%
-  select(group, ECO_NAME, ECO_NUM, WWF_MHTNAM) %>%
-  #' casting to polygon because some ecoregions are multipolygons, and we
-  #' we need a list of individual polygons for MRF smooths
-  #' turning off spherical geometry to prevent `st_make_valid()` from making
-  #' russia span the whole northern hemisphere by inverting it
-  st_cast('POLYGON') %>% # assigns attributes to sub-geometries by default
-  st_make_valid() %>% # to drop duplicate vertices
-  mutate(poly_id = paste('poly', 1:n()),
-         area_km2 = as.numeric(st_area(.)) / 1e6) %>%
-  filter(WWF_MHTNAM != 'Inland Water') # dropping 17 inland water polygons
-
-# drop points based on plot(n_data ~ area)
[…]

### 2025-05-16 analysis/004-merging-rasters.R
running aggregation by WWF realm

@@ -117,6 +117,9 @@ s_res <- 4 # spatial resolution
 t_res <- 3.5 # temporal resolution
 n_cells / s_res^2 / t_res / max_rows # modeling as a single dataset
 
+# non-integer time interval gives alternating repetition:
+table((0:20 - 0:20 %% t_res))
+
 # check cell sizes before and after spatial aggregation
 tibble(
   lat = seq(-90, 90, by = 30),
@@ -153,12 +156,54 @@ dates %>%
   group_by(n) %>%
   summarize(total = n())
 
-#' **HERE: need to decide s and t resoulutions, and split by WWF_REALM** 
-
 # need to aggregate temporally to reduce file size before saving ----
 # spatRast objects cannot be serialized (i.e., run in parallel):
 # https://stackoverflow.com/questions/67445883/terra-package-returns-error-when-try-to-run-parallel-operations/67449818#67449818
-# data for 2020 omly ----
+
+# calculate aggregated mean NDVI for each realm ---
+# cannot serialize rasters across cores, but can serialize realm names
+realms <- unique(ecoregions$WWF_REALM)
+
+plan(multisession(workers = min(availableCores() - 2, length(realms))))
+
+#' **RUNNING**
+future_map_chr(realms, function(.realm) {
+  shp <- filter(ecoregions, WWF_REALM == .realm) %>%
+    st_geometry() %>%
+    st_as_sf()
+  
+  dates %>%
+    filter(! is.na(file_name)) %>% # drop missing rasters
+    nest(cluster = ! group) %>%
+    mutate(central_date = map_dbl(cluster, function(.cl) {
+      mean(.cl$date)
+    }, .progress = 'Getting mean dates') %>%
+      as.Date(), # convert numeric back to date
+    n_rasters = map_int(cluster, nrow), # find number of rasters per group
+    # aggregate temporaly
+    ndvi_rast = map(cluster, function(.cl) {
+      map(.cl$file_name, \(.fn) rast(.fn, lyr = 'NDVI')) %>% # import
+        rast() %>% # convert list to stack of rasters
+        mask(shp) %>% # only keep the realm of interest
+        mean(na.rm = TRUE) %>% # aggregate temporally
+        return()
+    }, .progress = 'Calculating mean raster across time'),
+    # aggregating spatially
+    ndvi_rast = map(ndvi_rast, \(x) {
[…]
